---
title: "Supplementary material for *couplr: Optimal Matching with Verifiable Certificates and Sparse Edge Generation*"
author: "Gilles Colling"
geometry: a4paper,margin=1in
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: pdflatex
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{placeins}
  - \setlength{\tabcolsep}{4pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "100%")

d <- function(f) read.csv(file.path("..", "data", f), stringsAsFactors = FALSE)

regime_runs    <- d("regime-runs.csv")
regime_cells   <- d("regime-cells.csv")
regime_results <- d("regime-results.csv")
regime_verdict <- d("regime-verdict.csv")
igrid_runs     <- d("implicit-grid-runs.csv")
igrid          <- d("implicit-grid-results.csv")
memory         <- d("memory-results.csv")
imp            <- d("implicit-results.csv")
scaling_runs   <- d("scaling-runs.csv")
scaling_res    <- d("scaling-results.csv")
bench_tab      <- d("benchmark-table.csv")

kb <- function(x, caption, align = NULL, ...) {
  knitr::kable(x, booktabs = TRUE, longtable = TRUE, row.names = FALSE,
               align = align, caption = caption, ...)
}
## Compact enough not to wrap inside a table cell.
shape <- function(n_rows, n_cols) sprintf("%dx%d", n_rows, n_cols)
## The pattern labels are read off the cells the grid actually ran, so a
## pattern added to the benchmark appears here without being named twice.
pat_share <- tapply(regime_cells$finite_share, regime_cells$pattern, median)
pat_comp  <- tapply(regime_cells$components, regime_cells$pattern, max)
pat_short <- ifelse(names(pat_share) == "none", "complete",
              ifelse(pat_comp > 1,
                     sprintf("%.0f%%, %d blocks", 100 * pat_share, pat_comp),
                     sprintf("%.0f%%", 100 * pat_share)))
names(pat_short) <- names(pat_share)
short_pat <- function(x) unname(ifelse(is.na(pat_short[x]), x, pat_short[x]))
```

# What this file holds

The article reports summaries of four measurements that are too large to print
in full: the solver grid over problem regimes, the edge-generation loop over
covariate counts, point clouds, calipers and sizes, the peak memory of every
mode and comparator, and the raw runs behind the scaling table. This file
carries the definitions the measurements are stated over, the protocol they
were taken under, and summary and worst-case views of each one, and names the
script that produced each so any of them can be re-measured. It also carries the
parts of the software design the article points at: the object model, the C++
organisation and the dependency footprint, the proof of the sentinel-padding
lemma the article states, the glossary behind the solver figure's labels, and
the outcome analysis the article points at.

Every cell of every measurement is in the CSV files listed below, which ship
with the submission under `data/`. The tables here select from those files; they
do not replace them. Each section names the file that carries its cells in full.

```{r provenance}
env <- tryCatch(readLines(file.path("..", "data", "ENVIRONMENT.txt")),
                error = function(e) character(0))
## The host line runs past the width a verbatim block holds, so a line over
## that width is folded and its continuation indented to the value column.
if (length(env)) {
  fold <- function(x, width = 80) {
    if (nchar(x) <= width) return(x)
    lead <- attr(regexpr("^ *", x), "match.length")
    gap  <- attr(regexpr("^\\S+ +", substring(x, lead + 1)), "match.length")
    head_len <- lead + if (lead == 0 && gap > 0) gap else 0
    lines <- strwrap(substring(x, head_len + 1), width = width - head_len)
    paste(c(paste0(substring(x, 1, head_len), lines[1]),
            paste0(strrep(" ", head_len), lines[-1])), collapse = "\n")
  }
  cat(paste(vapply(env, fold, character(1)), collapse = "\n"))
}
```

| Section | Script | Files |
|---|---|---|
| Solver selection across regimes | `scripts/bench_regimes.R` | `regime-runs.csv`, `regime-cells.csv`, `regime-results.csv`, `regime-verdict.csv` |
| The edge-generation loop | `scripts/bench_implicit_grid.R` | `implicit-grid-runs.csv`, `implicit-grid-results.csv` |
| Peak memory | `scripts/bench_memory.R` | `memory-results.csv` |
| Representation against solver | `scripts/bench_implicit.R` | `implicit-results.csv` |
| Scaling | `scripts/bench_scaling.R` | `scaling-runs.csv`, `scaling-results.csv` |
| The solver figure | `scripts/make-figure.R` | `benchmark-table.csv` |

Every script is resume-safe: it writes its rows as it goes and skips a cell it
already holds, so a run that is interrupted continues where it stopped. Moving
the CSV aside forces a full re-measurement. `paper/run_bench_suite.sh` runs all
of them in order on one machine, which is how the numbers here were produced.

The scripts load couplr from a source checkout with `pkgload::load_all()` and
write their CSVs under `paper/`, so they run from a clone of the package
repository rather than from the submission archive:

```
git clone https://github.com/gcol33/couplr
cd couplr
sh paper/run_bench_suite.sh
```

The scaling, implicit, implicit-grid, regimes and memory scripts also take
`--quick`, which runs a reduced grid.

\FloatBarrier

# What a tolerance costs the conclusion

`verify_assignment()` decides its conditions in exact arithmetic by default, and
the article states the bound that applies when a tolerance is used instead. The
derivation is here.

A certificate at tolerance $\epsilon$ accepts potentials that violate dual
feasibility by $\epsilon$ on any pair, miss tightness by $\epsilon$ on any matched arc,
and leave $|v_j| \le \epsilon$ on any unused column. Write $n$ for the rows
and $m$ for the columns, with $n \le m$.

Shifting to $u_i - \epsilon$ and $v_j - \epsilon$ restores exact dual
feasibility and the sign condition, because every violated inequality was
violated by at most $\epsilon$. The shift costs the dual objective
$(n + m)\epsilon$, so the optimum is at least
$\sum_i u_i + \sum_j v_j - (n + m)\epsilon$.

Tightness within $\epsilon$ on the $n$ matched arcs, and $|v_j| \le \epsilon$ on
the $m - n$ unused columns, put the matching's own cost at most $m\epsilon$
above that same sum. Its excess over the optimum is therefore at most
$(n + 2m)\epsilon$.

At the largest shape in the article's benchmark, 16,667 rows against 33,333
columns, and the default $\epsilon = 10^{-9}$, that is
$8 \times 10^{-5}$ in the cost units.

The bound is why the band is not simply set to zero. The sign of a reduced cost
evaluated in double arithmetic can itself be wrong: with
$C_{ij} = \pm 2^{-60}$, $u_i = 1$ and $v_j = -1$, the first subtraction
rounds to $-1$ and the expression returns zero on a pair whose reduced cost is
not zero. In the negative case that is a dual infeasibility a zero-band check
would accept. The exact path splits each rounded addition into the sum and the
part the rounding lost, giving an expansion of doubles equal to the difference
in the reals, and takes the sign of that expansion [@Shewchuk1997]. A
rounding-error bound decides which pairs the double evaluation already settles,
so the expansion runs near tightness rather than over all $nm$ pairs.

# The integer conversion bit-scaling performs

`method = "gabow_tarjan"` runs on integers, so a real-valued cost matrix is
converted before it is solved and the conversion is part of the method.

Finite costs are shifted so the smallest is zero, multiplied by a scale factor
and rounded. The factor is set to keep the scaled costs clear of the sentinel
that marks a forbidden pair, and to keep the path sums the solver forms inside
64-bit arithmetic. A matrix whose range leaves no usable factor is refused, with
the limit named.

A matrix whose finite entries are each within $10^{-9}$ of an integer enters
unscaled, at the integer nearest each entry. Where those entries are integers
the conversion is exact and the optimum returned is the optimum of the instance
as supplied. Where they are merely near one, each cost moves by at most
$10^{-9}$, so the matching returned costs at most
$2\min(n,m) \times 10^{-9}$ more than that optimum.

Costs that are not integers are solved on the rounded instance. The matching
returned costs at most about $10^{-7}$ of the cost range more than the optimum
of the matrix as supplied at $n = 1000$, and $10^{-5}$ of it at
$n = 10{,}000$ on a square problem.

Potentials are divided by the factor and shifted back, so they belong to the
original matrix. Whether the matching is optimal for that matrix rather than
only for the rounded instance is a question `verify_assignment()` answers, and
it is the reason the article reports solver status and certificate separately.

# The bound the pricer prunes with

Where the metric admits a ball bound, the columns of a lazy cost source are
held in a metric tree and the pricing sweep discards a subtree without visiting
it.

For a subtree $S$ and row $i$ the test is
$$
L(i, S) - u_i - \max_{j \in S} v_j \ge 0 ,
$$
where $L(i, S)$ is a lower bound on $\min_{j \in S} C_{ij}$ read off the
node rather than the minimum itself. If the test holds, no column the node
carries can price in, and the subtree is discarded.

A node carries two summaries, because a distance and a caliper are stated in
different coordinates. The first is a centre and radius in whitened
coordinates, which is what $L(i, S)$ reads. The second is an axis-aligned
bounding box in the original covariates, because a caliper is a box there and a
whitened ball cannot exclude one; whitening is a rotation and a scaling, so it
destroys the axis alignment the per-variable window is stated in.

Mahalanobis distance is Euclidean after whitening, so one tree serves it and
the Euclidean metrics alike. Its cost is quadratic in the number of covariates,
dear enough to pay for the bound at every dimension measured, so it always
takes the tree. The metrics whose cost is linear in the covariates take it only
in low dimension. Manhattan and Chebyshev carry no ball bound, a covariance
with no Cholesky factor has no whitened coordinates, and a custom distance
function is opaque; each falls back to reading the pairs, which is the same
answer for more work rather than a different answer.

Every bound is lowered by an allowance before it is compared, so a subtree is
discarded only where it clears the threshold by more than the bound can be
wrong. The allowance covers the rounding of the centre distance and the radius,
each a sum of squares and a square root, and for Mahalanobis the residual of
the factorization the whitening uses, measured against the matrix it came from
rather than assumed. A subtree that does not clear the threshold is descended
into, which costs the evaluations the bound would have saved and not the edge,
so an allowance set too high loses time where one set too low would lose an
edge.

# What the restricted solve's optimum can cost

Proposition 1 is exact: if no omitted pair has a negative reduced cost, the
restricted optimum is optimal for the complete problem. The loop as it runs
weakens that on four counts, and they are worth naming separately rather than
folded into one term.

Write $n$ for the rows and $m$ for the columns.

- $\delta$, the duality gap the restricted master's own certificate reports.
  The restricted solution costs at most
  $\sum_i u_i + \sum_j v_j + \delta$. Every run in the article reported
  $\delta = 0$.
- $\epsilon$, the pricing threshold. A pair enters on
  $\bar{C}_{ij} < -\epsilon$, so termination establishes
  $\bar{C}_{ij} \ge -\epsilon$ over the omitted pairs rather than
  $\bar{C}_{ij} \ge 0$. Shifting the row potentials to
  $u_i - \epsilon$ restores dual feasibility over every admissible pair and
  costs the dual objective $n\epsilon$.
- The sign condition on the column potentials. Dual feasibility for a
  rectangular problem also asks $v_j \le 0$ on the columns an assignment may
  leave unused. The flow master returns potentials satisfying it by
  construction on the arcs it carries, and `verify_assignment()` checks it
  rather than assuming it; shifting $u$ alone does not repair a positive $v_j$.
- $\eta$, the allowance the pricing bound carries where a metric tree is
  used. A subtree is discarded only where its bound clears the threshold by
  more than $\eta$, so the statement termination establishes is
  $\bar{C}_{ij} \ge -\epsilon$ with the bound's own error already
  taken out rather than assumed away.

Taken together the complete optimum is at least
$\sum_i u_i + \sum_j v_j - n\epsilon$, and the returned matching's excess
over it is at most $\delta + n\epsilon$ under the first two, with the third
and fourth entering as conditions that must hold rather than as further slack:
the sign condition is checked, and the tree allowance is subtracted before any
prune fires.

At $\epsilon = 10^{-9}$ and the largest shape in the article's benchmark,
16,667 rows, $n\epsilon$ is $2 \times 10^{-5}$ in the cost units.

# The sentinel padding lemma

Tight constraints can leave the matching problem infeasible, and couplr answers
it by replacing the forbidden entries with a finite sentinel and re-solving.
The article states the lemma that makes a minimum-cost solve of the padded
problem return the lexicographic optimum, the largest admissible matching and
the cheapest among those of its size, and defers the proof here.

Write $[\ell, h]$ for the range of the admissible costs of the pruned
submatrix, $k$ for the smaller of its two dimensions and $\sigma$ for the
sentinel. Every feasible assignment of the padded problem matches all $k$ rows,
so it splits into $s$ sentinel edges and $k - s$ real ones and costs
$s\sigma + R$ with $R \in [(k-s)\ell, (k-s)h]$.

**Lemma 1.** *If $\sigma > k(|\ell| + |h|)$, then a padded assignment using more
sentinel edges than another costs strictly more.*

*Proof.* Let $M'$ and $M$ use $s'$ and $s$ sentinel edges with $s' > s$, and
write $R'$ and $R$ for their real costs. Then
$$
\text{cost}(M') - \text{cost}(M) = (s' - s)\sigma + R' - R
\ge \sigma + (k - s')\ell - (k - s)h
\ge \sigma - k|\ell| - k|h| > 0 . \qquad \square
$$

couplr takes $\sigma = (k+1)(|\ell| + |h|) + 1$, which satisfies the hypothesis
whatever the signs of $\ell$ and $h$ and when the admissible costs are all
zero, and leaves a margin of $|\ell| + |h| + 1$ between a matching and one
using more sentinel edges. `assignment(cardinality = "maximum")` prices its dummy
columns with the same quantity, and a maximisation is solved by negation with
the sentinel negated alongside it, so the lemma applies to the negated
instance. The reported objective is recomputed over the real pairs, so no
sentinel enters it.

The lemma's hypothesis is a statement about the numbers actually used, so it is
checkable on the doubles in hand and couplr's $\sigma$ satisfies it by
construction. What the lemma does not settle is the solve. A solver working
where a double no longer resolves a gap of $|\ell| + |h| + 1$ cannot be relied
on to order two assignments the padding separates by that much, so couplr
refuses a padded objective above $2^{53}$ rather than returning an ordering its
arithmetic may not support. That threshold is a precondition on the solve and
not a proof of it: $2^{53}$ bounds the integers a double carries exactly, and
the padded costs need not be integers. What establishes that a particular
padded solve reached its optimum is the certificate `verify_assignment()`
returns for the padded matrix, and the lemma carries that verdict across to the
lexicographic objective.

\FloatBarrier

# Solver glossary

The article's solver figure labels its traces with abbreviations short enough to
sit inside a panel. The mapping below gives, for each label, the value the
`method` argument of `assignment()` and `lap_solve()` takes, the family the
benchmark table groups it under, and the regime the solver is written for. The
labels and families are read from `benchmark-table.csv`, the same file the
figure is drawn from.

```{r solver-glossary, results="asis"}
solver_note <- c(
  jv              = "Column reduction and auction-style initialisation; the general-purpose default on a dense problem.",
  sap             = "Shortest augmenting path over the shared flow model; carries sparsity well.",
  sap_dense       = "Shortest augmenting path with a linear scan in place of a heap; a dense cost matrix.",
  lapmod          = "Sparse Jonker-Volgenant variant over an adjacency list; more than half the entries forbidden.",
  hungarian       = "The classical Hungarian method, O(n^3).",
  munkres         = "Matrix-form Kuhn-Munkres, O(n^4), kept as a reference implementation.",
  auction         = "Bertsekas auction with adaptive epsilon.",
  auction_gs      = "Gauss-Seidel auction variant; spatial structure.",
  auction_scaled  = "Epsilon-scaling auction; large dense problems.",
  csa             = "Goldberg-Kennedy cost scaling; medium to large problems.",
  gabow_tarjan    = "Gabow-Tarjan bit scaling with complementary slackness; integer costs.",
  ssap_bucket     = "Dial bucket priority queue over augmenting paths; integer costs.",
  csflow          = "Successive shortest paths with Johnson potentials on the flow model.",
  cycle_cancel    = "Cycle cancelling with Karp's minimum-mean cycle.",
  network_simplex = "Network simplex with a spanning-tree representation.",
  push_relabel    = "Goldberg-Tarjan cost-scaling push-relabel.",
  ramshaw_tarjan  = "Ramshaw-Tarjan; strongly rectangular problems, where padding to a square would dominate.",
  hk01            = "Hopcroft-Karp cardinality matching; binary or constant finite costs.",
  bruteforce      = "Exact enumeration; at most eight rows and eight columns.",
  auto            = "The dispatch rules, timed as a solver of its own so the probing pass is inside its number.")

gl <- unique(bench_tab[, c("method", "label", "family")])
gl <- gl[order(gl$family != "Auto", gl$family, gl$method), ]
gl$Meaning <- unname(solver_note[gl$method])
## A pipe table rather than kable: the meaning column has to wrap, which a
## latex tabular built from l columns cannot do.
cat("| Label | `method =` | Family | Meaning |\n")
cat("|:------------|:--------------|:---------------|:----------------------------------------|\n")
cat(sprintf("| %s | `%s` | %s | %s |\n", gl$label, gl$method, gl$family,
            gl$Meaning), sep = "")
cat("\nTable: Plot label, argument value, family and intended regime for every solver the article's figure draws.\n\n")
```

Three further routines are reachable but not drawn in the figure: the
bottleneck objective, which minimises the largest pair cost instead of the sum,
Murty's ranking procedure for the k cheapest assignments, and Sinkhorn
iteration for the entropy-regularised relaxation.

\FloatBarrier

# From matched pairs to an estimate

The article shows the matched sample and defers the outcome analysis here. The
match is rebuilt below so this file stands on its own.

```{r estimate-setup, echo=TRUE, message=FALSE, warning=FALSE}
library(couplr)
data(hospital_staff)
treated <- transform(hospital_staff$nurses_extended,   id = nurse_id)
control <- transform(hospital_staff$controls_extended, id = nurse_id)
covars  <- c("age", "experience_years", "certification_level")
m <- match_couples(treated, control, vars = covars, auto_scale = TRUE)
```

Taking the hourly rate as the outcome, the paired difference within each
matched pair estimates the association between full-time status and pay among
nurses comparable on age, experience and certification.

```{r estimate-effect, echo=TRUE}
matched <- join_matched(m, treated, control)
d  <- matched$hourly_rate_left - matched$hourly_rate_right
tt <- t.test(d)
c(estimate = mean(d), lower = tt$conf.int[1], upper = tt$conf.int[2])
```

Full-time nurses earn `r sprintf("%.2f", mean(matched$hourly_rate_left - matched$hourly_rate_right))`
more per hour than their matched part-time counterparts, with a 95 percent
interval that excludes zero. As a paired difference on a matched sample this
estimates an adjusted association. Reading it as the causal effect of full-time
status is what rests on assumptions no matching method can verify: conditional
ignorability, positivity, and the stable unit treatment value assumption
(SUTVA), under which a unit's outcome depends on its own treatment alone. The
matched output can also be passed to an outcome model for regression
adjustment; valid estimation and uncertainty quantification still depend on the
estimator used.

What matching can offer against the first of those assumptions is a statement
about how strong an unmeasured confounder would have to be to overturn the
result. `sensitivity_analysis()` implements Rosenbaum bounds (Rosenbaum 2002)
for the Wilcoxon signed-rank statistic on one-to-one matched pairs, reporting
upper and lower p-value bounds across a grid of the odds-ratio parameter
$\Gamma$.

```{r estimate-sensitivity, echo=TRUE}
sens <- sensitivity_analysis(m, treated, control, outcome_var = "hourly_rate")
sens$results[sens$results$gamma %in% c(1, 1.5, 1.75, 2), ]
```

At $\Gamma = 1$, meaning no hidden bias, the association is significant. It
survives an unmeasured confounder that raises one pair member's odds of being
full-time by half, and stops being significant at
$\Gamma =$ `r sens$critical_gamma`. A reader can judge from that number whether
a confounder of that strength is plausible in this setting; the analysis
quantifies exposure to hidden bias rather than removing it.

\FloatBarrier

# Solver selection across problem regimes

`assignment(method = "auto")` chooses a solver from five rules read off one pass
over the cost matrix: enumerate an at most 8 by 8 problem, use a cardinality
algorithm where the finite costs are constant or binary, use the sparse
Jonker-Volgenant variant where more than half the entries are forbidden, use the
shortest-augmenting-path solver where there are at least three times as many
columns as rows, and otherwise use Jonker-Volgenant. The article's solver figure
varies problem size on square dense integer costs, which is the last rule's
regime and no other. This grid varies the properties the rules are stated over.

A cell is a cost regime, a forbidden pattern and a shape. Each cell is generated
as several independent instances, each instance is timed several times, and
every solver in the panel sees the same instance. The timing a cell reports for
a solver is the median across its instances, where each instance contributes the
median of its own repetitions, so the spread is between problems rather than
between repetitions of one problem. A solver that errored or timed out on a cell
has no row there.

```{r regime-design}
reg_lab <- c(int_uniform = "integer, uniform on 1..10,000",
             dbl_uniform = "double, uniform on (0, 1)",
             binary = "binary, 0 or 1",
             constant = "constant",
             tied = "integer, five distinct values",
             heavy_tailed = "double, lognormal",
             metric_uniform = "euclidean over uniform points in 8 dimensions",
             metric_clustered = "euclidean over eight clusters in 8 dimensions")
pat_tab <- data.frame(
  Level = names(pat_share),
  Meaning = sprintf("%s admissible, %s",
                    ifelse(names(pat_share) == "none", "every pair",
                           sprintf("%.0f percent of pairs", 100 * pat_share)),
                    ifelse(pat_comp > 1,
                           sprintf("in %d disconnected components", pat_comp),
                           "unstructured")))
kb(data.frame(Factor = c(rep("cost regime", length(reg_lab)),
                         rep("forbidden pattern", nrow(pat_tab))),
              Level = c(names(reg_lab), pat_tab$Level),
              Meaning = c(unname(reg_lab), pat_tab$Meaning)),
   caption = "The factors the solver grid crosses. The admissible shares are the medians of the shares the generated cells actually held. Shapes and sizes are in the next table.",
   align = "lll")
```

```{r regime-shapes}
sh <- unique(regime_cells[, c("tier", "n_rows", "n_cols")])
sh <- sh[order(sh$tier, sh$n_rows, sh$n_cols), ]
sh$aspect <- sprintf("1:%.0f", sh$n_cols / sh$n_rows)
sh$Shape <- shape(sh$n_rows, sh$n_cols)
tier_panel <- tapply(regime_runs$method, regime_runs$tier,
                     function(x) length(unique(x)))
sh$`Solvers in the panel` <- as.integer(tier_panel[sh$tier])
kb(sh[, c("tier", "Shape", "aspect", "Solvers in the panel")],
   caption = "Shapes. The base tier runs the whole panel at a size every solver in it reaches; the large tier repeats the crossing at a size only the scalable solvers reach.",
   align = "llrr")
```

## What the rules cost

For each cell, the solver `"auto"` selected is timed as a solver of its own, so
its number carries the probing pass the rules cost, and it is compared against
the fastest named solver in the same cell. Ratio below is the time the
dispatched solver took over the time the cell's fastest named solver took, so
1.00x means the rule picked the best available solver and 2.00x means it cost
twice the best.

```{r regime-by-rule}
by_rule <- do.call(rbind, lapply(split(regime_verdict, regime_verdict$auto_rule),
                                 function(g) data.frame(
  Rule = g$auto_rule[1],
  Cells = nrow(g),
  `Picked the cell's fastest` = sum(g$auto_picks == g$best_method),
  `Median ratio` = sprintf("%.2fx", median(g$ratio)),
  `Worst ratio` = sprintf("%.2fx", max(g$ratio)),
  check.names = FALSE)))
kb(by_rule[order(-by_rule$Cells), ],
   caption = "Each dispatch rule, over the cells where it fired.",
   align = "lrrrr")
```

```{r regime-by-regime}
grp <- function(f) {
  do.call(rbind, lapply(split(regime_verdict, regime_verdict[[f]]),
                        function(g) data.frame(
    Level = as.character(g[[f]][1]),
    Cells = nrow(g),
    `Picked the cell's fastest` = sum(g$auto_picks == g$best_method),
    `Median ratio` = sprintf("%.2fx", median(g$ratio)),
    `Worst ratio` = sprintf("%.2fx", max(g$ratio)),
    check.names = FALSE)))
}
br <- grp("regime")
bp <- grp("pattern"); bp$Level <- short_pat(bp$Level)
bt <- grp("tier")
kb(rbind(cbind(Factor = "cost regime", br),
         cbind(Factor = "forbidden pattern", bp),
         cbind(Factor = "tier", bt)),
   caption = "The same comparison collapsed over each factor of the grid in turn.",
   align = "llrrrr")
```

```{r regime-worst}
w <- regime_verdict[order(-regime_verdict$ratio), ]
w <- head(w, 15)
kb(data.frame(Regime = w$regime, Pattern = short_pat(w$pattern),
              Shape = gsub(" ", "", w$shape),
              Picked = w$auto_picks,
              `Best` = w$best_method,
              `Auto (ms)` = sprintf("%.3g", 1000 * w$auto_s),
              `Best (ms)` = sprintf("%.3g", 1000 * w$best_s),
              Ratio = sprintf("%.2fx", w$ratio), check.names = FALSE),
   caption = "The 15 cells where the dispatched solver was furthest off the cell's fastest.",
   align = "lllllrrr")
```

The rule, the solver it picked, the fastest solver and the ratio between them
are in `regime-verdict.csv` for all `r nrow(regime_verdict)` cells.

## Every solver against the fastest in its cell

The table below reduces each solver to its position across the grid. A solver's
ratio in a cell is its median time over the median time of the fastest solver in
that cell, so the reduction is within-cell and a solver that entered only the
easy cells is not thereby favoured. Solvers differ in how many cells they
entered, since the large tier runs a reduced panel and a solver that errored or
timed out on a cell has no row there.

```{r regime-per-solver}
key <- paste(regime_results$tier, regime_results$regime, regime_results$pattern,
             regime_results$n_rows, regime_results$n_cols)
best_in_cell <- tapply(regime_results$median_s, key, min)
regime_results$ratio <- regime_results$median_s / best_in_cell[key]
ps <- do.call(rbind, lapply(split(regime_results, regime_results$method),
                            function(g) data.frame(
  Solver = g$method[1],
  `Cells entered` = nrow(g),
  `Times fastest` = sum(g$ratio <= 1 + 1e-12),
  `Median ratio` = sprintf("%.2fx", median(g$ratio)),
  `Q75 ratio` = sprintf("%.2fx", quantile(g$ratio, 0.75)),
  `Worst ratio` = sprintf("%.1fx", max(g$ratio)),
  `Fastest cell (ms)` = sprintf("%.3g", 1000 * min(g$median_s)),
  `Slowest cell (ms)` = sprintf("%.3g", 1000 * max(g$median_s)),
  check.names = FALSE)))
ps <- ps[order(as.numeric(sub("x", "", ps$`Median ratio`))), ]
kb(ps, caption = "Every solver over the cells it entered, as a ratio to the fastest solver in the same cell.",
   align = "lrrrrrrr")
```

```{r regime-pointer, results="asis"}
cells <- length(unique(key))
cat(sprintf("The per-cell medians and quartiles behind this table are in `regime-results.csv`, %d rows covering %d cells and %d solvers, and the individual timed repetitions are in `regime-runs.csv`, %d rows.\n\n",
            nrow(regime_results), cells, length(unique(regime_results$method)),
            nrow(regime_runs)))
```

## Do the solvers agree

Every solver in a cell solves the same instance, so their totals have to agree.
Each run below is compared against the median total of its own cell, and a run
counts as off that median where the difference exceeds $10^{-6}$ of it.

```{r regime-agreement, results="asis"}
ok <- regime_runs[regime_runs$status == "ok" & !is.na(regime_runs$total_cost), ]
akey <- paste(ok$tier, ok$regime, ok$pattern, ok$n_rows, ok$n_cols, ok$instance)
ok$ref <- ave(ok$total_cost, akey, FUN = median)
ok$rel <- abs(ok$total_cost - ok$ref) / pmax(1, abs(ok$ref))
n_solv <- length(unique(ok$method))
per_inst <- tapply(ok$method, akey, function(x) length(unique(x)))
off <- ok[ok$rel > 1e-6, ]
agree <- setdiff(unique(ok$method), unique(off$method))
cat(sprintf("%d instances carry %d to %d solvers each. %d of the %d solvers return the cell median exactly on every run. The other %d are below.\n\n",
            length(per_inst), min(per_inst), max(per_inst),
            length(agree), n_solv, length(unique(off$method))))
```

```{r regime-agreement-table}
if (nrow(off)) {
  os <- do.call(rbind, lapply(split(off, off$method), function(g) data.frame(
    Solver = g$method[1],
    `Runs` = sum(ok$method == g$method[1]),
    `Runs off the median` = nrow(g),
    `Regimes` = paste(sort(unique(g$regime)), collapse = ", "),
    `Worst relative difference` = sprintf("%.3g", max(g$rel)),
    check.names = FALSE)))
  kb(os[order(-os$`Runs off the median`), ],
     caption = "The solvers whose total differed from the median total of a cell they ran in.",
     align = "lrrlr")
}
```

Both are scaling methods, and both deviate on the lognormal regime alone, so
the rounding a scaling step applies over a cost range that wide may be what
separates them from the rest. The size of the two deviations differs by orders
of magnitude, and neither is a property of the regime: the remaining solvers
return one common total on every cell of it. Whether a particular matching is
optimal for the matrix as supplied is what `verify_assignment()` answers.

\FloatBarrier

# The edge-generation loop away from one cloud

The article's implicit results come from one cloud: eight independent normal
covariates with the treated group shifted, one draw per size. This grid sweeps
one factor at a time around that base, several seeds deep. Where the dense path
can afford the problem, the loop's answer is compared against it unit by unit.

```{r igrid-summary}
## Each sweep moves one factor, so a cell is named by the value that factor
## took; the rest of the key holds at the base and is stated in the caption.
sweep_level <- function(x) {
  ifelse(x$sweep == "caliper", as.character(x$caliper),
  ifelse(x$sweep == "cloud", x$cloud,
  ifelse(x$sweep == "dimension", sprintf("%d dimensions", x$dim),
         paste(formatC(x$n_total, big.mark = ",", format = "d"), "units"))))
}
g <- igrid[order(igrid$sweep, igrid$n_total, igrid$dim, igrid$cloud,
                 igrid$caliper), ]
kb(data.frame(Sweep = g$sweep, Level = sweep_level(g),
              Seeds = g$seeds,
              Rounds = sprintf("%g-%g", g$rounds_min, g$rounds_max),
              `Graph %` = sprintf("%.3f", g$graph_pct_med),
              `Distances` = sprintf("%.2fx", g$distances_x_med),
              `Implicit` = sprintf("%.3g s", g$implicit_s_med),
              `Lazy` = sprintf("%.3g s", g$lazy_s_med),
              `Speedup` = sprintf("%.2fx", g$speedup_med),
              check.names = FALSE),
   caption = "The loop by cell. A sweep moves one factor and holds the rest at the base cell, a gaussian cloud of 10,000 units in 8 dimensions with no caliper. Rounds is the range over seeds. Graph is the median share of the complete pair count the candidate set ended up holding, in percent. Distances is the median number of distance evaluations over all rounds as a multiple of the complete pair count. Implicit and lazy are median seconds over seeds and speedup is their median ratio.",
   align = "llrrrrrrr")
```

```{r igrid-certificate}
kb(data.frame(Sweep = g$sweep, Level = sweep_level(g),
              `Worst duality gap` = format(g$worst_gap, digits = 3),
              `All certified` = g$all_certified,
              `Equals the dense solve` = ifelse(is.na(g$equal_to_dense),
                                                "not run", g$equal_to_dense),
              check.names = FALSE),
   caption = "The certificate by cell. The duality gap is the residual the loop stopped on. Equals the dense solve compares the pairing unit by unit where the dense path could be run, and reads not run where the problem was too large for it.",
   align = "llrrr")
```

The medians above are taken over seeds, and the seed-to-seed spread is what
decides whether they carry. The next table reports, for each cell, the widest
separation the seeds produced.

```{r igrid-spread}
imp_runs <- igrid_runs[igrid_runs$memory_mode == "implicit" &
                         igrid_runs$status == "ok", ]
imp_runs$graph_pct <- 100 * imp_runs$candidate_edges / imp_runs$possible_edges
skey <- paste(imp_runs$sweep, imp_runs$cloud, imp_runs$dim, imp_runs$n_total,
              imp_runs$caliper)
sp <- do.call(rbind, lapply(split(imp_runs, skey), function(g2) data.frame(
  Sweep = g2$sweep[1], Level = sweep_level(g2[1, ]),
  ord_n = g2$n_total[1], ord_d = g2$dim[1], ord_c = g2$caliper[1],
  Seeds = nrow(g2),
  `Rounds` = sprintf("%g-%g", min(g2$n_rounds), max(g2$n_rounds)),
  `Seed width` = sprintf("%g-%g", min(g2$seed_width), max(g2$seed_width)),
  `Graph %` = sprintf("%.3f-%.3f", min(g2$graph_pct), max(g2$graph_pct)),
  `Seconds` = sprintf("%.3g-%.3g", min(g2$elapsed_s), max(g2$elapsed_s)),
  `Slowest / fastest` = sprintf("%.2fx", max(g2$elapsed_s) / min(g2$elapsed_s)),
  check.names = FALSE)))
sp <- sp[order(sp$Sweep, sp$ord_n, sp$ord_d, sp$ord_c, sp$Level), ]
sp <- sp[, setdiff(names(sp), c("ord_n", "ord_d", "ord_c"))]
kb(sp, caption = "Across the seeds of a cell: the range of each quantity and the ratio of the slowest run to the fastest. Seed width is the number of columns the first round gave each row.",
   align = "llrrrrrr")
```

```{r igrid-failures, results="asis"}
bad <- igrid_runs[igrid_runs$status != "ok", ]
if (nrow(bad)) {
  kb(data.frame(Sweep = bad$sweep, Cloud = bad$cloud, Dim = bad$dim,
                Units = bad$n_total, Caliper = bad$caliper,
                Mode = bad$memory_mode, Seed = bad$seed_index,
                Status = bad$status, check.names = FALSE),
     caption = "Cells that did not finish.", align = "llllllll")
} else {
  cat(sprintf("Every cell of the grid finished: %d runs over %d cells and %d memory modes, all with status ok.\n\n",
              nrow(igrid_runs), nrow(igrid),
              length(unique(igrid_runs$memory_mode))))
}
```

```{r igrid-pointer, results="asis"}
cat(sprintf("Each of the %d runs behind these tables is a row of `implicit-grid-runs.csv`, carrying its seed, mode, timing, round count, candidate and evaluated edge counts, duality gap and comparison against the dense solve. The per-cell medians are in `implicit-grid-results.csv`.\n\n",
            nrow(igrid_runs)))
```

\FloatBarrier

# Peak memory

Peak resident set size is a property of a process, so each cell here is a fresh
R process that loads what it needs, does one matching and exits, measured from
outside by the platform's `time` utility. The baseline row of a size is a
process that loaded the same packages and matched nothing, and it is what the
increment column is measured against. `dense_matrix` builds the distance matrix
and stops, which separates the representation from the solver that runs on it.

```{r memory-table}
mm <- memory[order(memory$n_total, memory$arm), ]
kb(data.frame(Units = formatC(mm$n_total, big.mark = ",", format = "d"),
              Arm = mm$arm,
              `Peak RSS` = sprintf("%.0f", mm$peak_rss_mb),
              `Over baseline` = ifelse(is.na(mm$over_baseline_mb), "-",
                                       sprintf("%.0f", mm$over_baseline_mb)),
              `R heap peak` = sprintf("%.0f", mm$heap_peak_mb),
              `Seconds` = sprintf("%.3g", mm$elapsed_s),
              Outcome = sub(":.*$", "", mm$status), check.names = FALSE),
   caption = "Peak resident set size by arm and problem size, in megabytes.",
   align = "llrrrrl")
```

```{r memory-heuristic}
dm <- memory[memory$arm == "dense_matrix" & !is.na(memory$over_baseline_mb), ]
if (nrow(dm)) {
  kb(data.frame(Units = formatC(dm$n_total, big.mark = ",", format = "d"),
                `Matrix` = sprintf("%.0f", dm$matrix_mb),
                `Measured over baseline` = sprintf("%.0f", dm$over_baseline_mb),
                `Measured / matrix` = sprintf("%.2f", dm$over_baseline_mb / dm$matrix_mb),
                `Package estimate` = sprintf("%.0f", dm$estimate_mb),
                check.names = FALSE),
     caption = "The four-fold estimate against the measurement, in megabytes. Matrix is the bytes the dense cost matrix itself needs at that shape, and package estimate is what estimate_dense_matrix_mb() predicts, which is four times the matrix and is what the memory dispatcher warns from. The estimate is deliberately conservative: it exists to avoid under-warning before a large allocation, not to predict the peak.",
     align = "lrrrr")
}
```

```{r memory-pointer, results="asis"}
mb <- memory[memory$status != "ok", ]
if (nrow(mb)) {
  cat(sprintf("The %s arm at %s units did not finish, with `%s`. ",
              mb$arm[1], formatC(mb$n_total[1], big.mark = ",", format = "d"),
              mb$status[1]))
}
cat(sprintf("All %d measured processes are in `memory-results.csv`, which also carries the matrix and estimate columns for every arm.\n\n",
            nrow(memory)))
```

\FloatBarrier

# Representation against solver

The implicit mode is a representation and a solver at once: its restricted
master is the flow model, and naming a solver under it is an error rather than a
setting. Reading its timing as a representation comparison would therefore
attribute the flow solver's behaviour to the representation. The table below is
the decomposition: down a column, one solver over two representations; across a
row, one representation under two solvers.

```{r repr-solver}
sizes <- sort(unique(imp$n_total))
cell <- function(mode, n) {
  v <- imp$elapsed_s[imp$memory_mode == mode & imp$n_total == n]
  if (!length(v) || is.na(v[1])) return("not run")
  if (v[1] < 1) sprintf("%.0f ms", 1000 * v[1]) else sprintf("%.3g s", v[1])
}
kb(data.frame(
  Units = format(sizes, big.mark = ","),
  `Dense, JV` = vapply(sizes, function(n) cell("dense", n), character(1)),
  `Lazy, JV` = vapply(sizes, function(n) cell("lazy", n), character(1)),
  `Dense, flow` = vapply(sizes, function(n) cell("dense_csflow", n), character(1)),
  `Implicit` = vapply(sizes, function(n) cell("implicit", n), character(1)),
  check.names = FALSE),
  caption = "Wall-clock seconds by representation and solver. Dense and lazy differ only in whether the matrix is materialised; dense and dense-flow differ only in the solver; implicit is the mode as a user meets it.",
  align = "lrrrr")
```

The certificate, pair count, candidate edge count and round count of each of
these runs are in `implicit-results.csv`.

\FloatBarrier

# Scaling

The scaling table in the article reports the median across instances. The
dispersion behind those medians is below: an instance is an independently
generated problem of that size, a repetition is one timed solve of that problem,
and every package sees the same instances.

```{r scaling-summary}
sres <- scaling_res[order(scaling_res$n_total, scaling_res$package), ]
fin <- !is.na(sres$median_s)
num <- function(x, fmt = "%.4g") ifelse(is.na(x), "-", sprintf(fmt, x))
kb(data.frame(Units = format(sres$n_total, big.mark = ","),
              Package = sres$package,
              Instances = sres$instances, Runs = sres$runs,
              `Median (s)` = num(sres$median_s),
              `Min (s)` = num(sres$min_s),
              `Max (s)` = num(sres$max_s),
              `Max / min` = ifelse(fin & sres$min_s > 0,
                                   sprintf("%.2fx", sres$max_s / sres$min_s), "-"),
              Outcome = sub(":.*$", "", sres$status),
              check.names = FALSE),
   caption = "Every size and package of the scaling benchmark. Max over min is the spread across all repetitions of that cell, so it carries both the between-instance and the within-instance variation.",
   align = "llrrrrrrl")
```

```{r scaling-failures}
sb <- scaling_runs[scaling_runs$status != "ok", ]
if (nrow(sb)) {
  agg <- as.data.frame(table(sb$n_total, sb$package, sb$status),
                       stringsAsFactors = FALSE)
  agg <- agg[agg$Freq > 0, ]
  names(agg) <- c("Units", "Package", "Outcome", "Runs")
  agg <- agg[order(as.numeric(agg$Units), agg$Package), ]
  agg$Units <- formatC(as.numeric(agg$Units), big.mark = ",", format = "d")
  kb(agg[, c("Units", "Package", "Runs", "Outcome")],
     caption = "The runs that did not return a time.", align = "llrl")
}
```

```{r scaling-pointer, results="asis"}
cat(sprintf("Each of the %d individual repetitions, with its instance index, seed, seconds and status, is a row of `scaling-runs.csv`. The per-cell quartiles are in `scaling-results.csv`.\n\n",
            nrow(scaling_runs)))
```

\FloatBarrier

# Software design

The article describes the three layers the package is organised in, the dispatch
rules, the memory modes and how the package is verified. This section carries the
object model, the C++ organisation and the dependency footprint.

## Object model

The object model is S3 throughout. Matching results carry the class vector
`c("matching_result", "couplr_result")`, with the design-specific classes
(`full_matching_result`, `cem_result`, `subclass_result`) sharing the
`couplr_result` parent, which is what lets `balance_diagnostics()`,
`join_matched()`, `match_data()` and the `bal.tab()` methods dispatch on it once
rather than being rewritten per design. Results are plain lists of tibbles, so a
user who wants something the package does not provide can reach in and take it.
`lap_solve()` returns a tibble subclassed as `lap_solve_result` with `source`,
`target` and `cost` columns, so an optimum is usable in a plot or a join without
an extraction step, while `get_total_cost()` and `get_method_used()` recover the
metadata carried in attributes.

## C++ organisation

All nineteen solvers are written from scratch in C++ and exposed through Rcpp,
with `src/core` holding the cost-source types and shared utilities with no Rcpp
dependency, `src/solvers` one algorithm per file with a thin `_rcpp.cpp` wrapper
each, `src/flow` the flow model, its two solvers and the edge-generation loop,
and `src/interface` the conversion between R objects and the internal types. The
separation means the solver bodies are ordinary C++ that can be compiled and
tested outside R, which is what the C++ test suite does.

Solvers are templated on their cost source rather than written against a concrete
matrix. A cost source provides `at(i, j)`, `allowed(i, j)` and `empty()`, and two
types satisfy that interface: `CostMatrix`, which owns a dense flat array plus a
mask, and `LazyCostMatrix`, which owns the two feature matrices and computes
$C_{ij}$ on demand from the requested metric, applying calipers and the distance
ceiling in `allowed()`. A handful of hot loops index the dense mask array
directly for speed, which the lazy type cannot support; these are guarded by a
`supports_raw_mask` trait and an `if constexpr` branch rather than by duplicating
the solver. One solver body therefore serves both cost representations, and the
pricing oracle of the edge-generation loop needs `at(i, j)` and `allowed(i, j)`
and nothing else, so it consumes the same concept and adds no third
representation of the costs.

`LazyCostMatrix` also bakes in the transformations that the dense path applies in
a separate preparation pass, namely mapping forbidden entries to a large sentinel
and negating costs when maximising. Doing this at construction is what allows the
untouched solver body to work on either type.

## Dependency footprint

couplr declares Rcpp, tibble, dplyr, rlang, purrr and `methods` in `Imports`, and
`Rcpp` alone in `LinkingTo`. The recursive closure of hard dependencies is 17
non-base packages, so a default installation brings 18 packages including couplr
itself. The comparable figures on the same day were 18 for optmatch, 8 for
MatchIt and 7 for designmatch.

Two choices sit behind that number. First, we use no external solver: all
nineteen algorithms are package source, so there is no dependency on an LP or MIP
library and no system solver to install. That choice also keeps `LinkingTo` to
`Rcpp` alone, which matters because every entry there forces a compile of that
package before couplr can be built from source; couplr declares only what its
headers actually include. Second, everything that serves a single feature lives
in `Suggests` and is loaded at call time. The animation and image-morphing
functions, the plotting helpers, the interoperability layer for MatchIt, optmatch
and cobalt, and the parallel back end are all optional in this sense, so a user
who only wants to match pays for none of them.
