---
title: "Supplementary material for *couplr: Optimal Matching with Verifiable Certificates and Sparse Edge Generation*"
author: "Gilles Colling"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: pdflatex
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "100%")

d <- function(f) read.csv(file.path("..", "data", f), stringsAsFactors = FALSE)

regime_runs    <- d("regime-runs.csv")
regime_cells   <- d("regime-cells.csv")
regime_results <- d("regime-results.csv")
regime_verdict <- d("regime-verdict.csv")
igrid_runs     <- d("implicit-grid-runs.csv")
igrid          <- d("implicit-grid-results.csv")
memory         <- d("memory-results.csv")
imp            <- d("implicit-results.csv")
scaling_runs   <- d("scaling-runs.csv")

kb <- function(x, caption, align = NULL, ...) {
  knitr::kable(x, booktabs = TRUE, longtable = TRUE, row.names = FALSE,
               align = align, caption = caption, ...)
}
## Compact enough not to wrap inside a table cell.
shape <- function(n_rows, n_cols) sprintf("%dx%d", n_rows, n_cols)
pat_short <- c(none = "complete", random_60 = "60%", random_25 = "25%",
               block_4 = "4 blocks")
short_pat <- function(x) unname(ifelse(is.na(pat_short[x]), x, pat_short[x]))
```

# What this file holds

The article reports summaries of four measurements that are too large to print
in full: the solver grid over problem regimes, the edge-generation loop over
covariate counts, point clouds, calipers and sizes, the peak memory of every
mode and comparator, and the raw runs behind the scaling table. This file
carries them in full, and names the script that produced each one so any of them
can be re-measured. It also carries the parts of the software design the article
points at: the object model, the C++ organisation and the dependency
footprint, the proof of the sentinel-padding lemma the article states, and
the outcome analysis the article points at.

```{r provenance}
env <- tryCatch(readLines(file.path("..", "data", "ENVIRONMENT.txt")),
                error = function(e) character(0))
if (length(env)) cat(paste(env, collapse = "\n"))
```

| Section | Script | Files |
|---|---|---|
| Solver selection across regimes | `scripts/bench_regimes.R` | `regime-runs.csv`, `regime-cells.csv`, `regime-results.csv`, `regime-verdict.csv` |
| The edge-generation loop | `scripts/bench_implicit_grid.R` | `implicit-grid-runs.csv`, `implicit-grid-results.csv` |
| Peak memory | `scripts/bench_memory.R` | `memory-results.csv` |
| Representation against solver | `scripts/bench_implicit.R` | `implicit-results.csv` |
| Scaling, every run | `scripts/bench_scaling.R` | `scaling-runs.csv`, `scaling-results.csv` |

Every script is resume-safe: it writes its rows as it goes and skips a cell it
already holds, so a run that is interrupted continues where it stopped. Moving
the CSV aside forces a full re-measurement. `paper/run_bench_suite.sh` runs all
of them in order on one machine, which is how the numbers here were produced.

The scripts load couplr from a source checkout with `pkgload::load_all()` and
write their CSVs under `paper/`, so they run from a clone of the package
repository rather than from the submission archive:

```
git clone https://github.com/gcol33/couplr
cd couplr
sh paper/run_bench_suite.sh
```

The scaling, implicit, implicit-grid, regimes and memory scripts also take
`--quick`, which runs a reduced grid.

\FloatBarrier

# The sentinel padding lemma

Tight constraints can leave the matching problem infeasible, and couplr answers
it by replacing the forbidden entries with a finite sentinel and re-solving.
The article states the lemma that makes a minimum-cost solve of the padded
problem return the lexicographic optimum, the largest admissible matching and
the cheapest among those of its size, and defers the proof here.

Write $[\ell, h]$ for the range of the admissible costs of the pruned
submatrix, $k$ for the smaller of its two dimensions and $\sigma$ for the
sentinel. Every feasible assignment of the padded problem matches all $k$ rows,
so it splits into $s$ sentinel edges and $k - s$ real ones and costs
$s\sigma + R$ with $R \in [(k-s)\ell, (k-s)h]$.

**Lemma 1.** *If $\sigma > k(|\ell| + |h|)$, then a padded assignment using more
sentinel edges than another costs strictly more.*

*Proof.* Let $M'$ and $M$ use $s'$ and $s$ sentinel edges with $s' > s$, and
write $R'$ and $R$ for their real costs. Then
$$
\text{cost}(M') - \text{cost}(M) = (s' - s)\sigma + R' - R
\ge \sigma + (k - s')\ell - (k - s)h
\ge \sigma - k|\ell| - k|h| > 0 . \qquad \square
$$

couplr takes $\sigma = (k+1)(|\ell| + |h|) + 1$, which satisfies the hypothesis
whatever the signs of $\ell$ and $h$ and when the admissible costs are all
zero, and leaves a margin of $|\ell| + |h| + 1$ between a matching and one
using more sentinel edges. `assignment(cardinality = "maximum")` prices its dummy
columns with the same quantity, and a maximisation is solved by negation with
the sentinel negated alongside it, so the lemma applies to the negated
instance. The reported objective is recomputed over the real pairs, so no
sentinel enters it.

The lemma's hypothesis is a statement about the numbers actually used, so it is
checkable on the doubles in hand and couplr's $\sigma$ satisfies it by
construction. What the lemma does not settle is the solve. A solver working
where a double no longer resolves a gap of $|\ell| + |h| + 1$ cannot be relied
on to order two assignments the padding separates by that much, so couplr
refuses a padded objective above $2^{53}$ rather than returning an ordering its
arithmetic may not support. That threshold is a precondition on the solve and
not a proof of it: $2^{53}$ bounds the integers a double carries exactly, and
the padded costs need not be integers. What establishes that a particular
padded solve reached its optimum is the certificate `verify_assignment()`
returns for the padded matrix, and the lemma carries that verdict across to the
lexicographic objective.

\FloatBarrier

# From matched pairs to an estimate

The article shows the matched sample and defers the outcome analysis here. The
match is rebuilt below so this file stands on its own.

```{r estimate-setup, echo=TRUE, message=FALSE, warning=FALSE}
library(couplr)
data(hospital_staff)
treated <- transform(hospital_staff$nurses_extended,   id = nurse_id)
control <- transform(hospital_staff$controls_extended, id = nurse_id)
covars  <- c("age", "experience_years", "certification_level")
m <- match_couples(treated, control, vars = covars, auto_scale = TRUE)
```

Taking the hourly rate as the outcome, the paired difference within each
matched pair estimates the association between full-time status and pay among
nurses comparable on age, experience and certification.

```{r estimate-effect, echo=TRUE}
matched <- join_matched(m, treated, control)
d  <- matched$hourly_rate_left - matched$hourly_rate_right
tt <- t.test(d)
c(estimate = mean(d), lower = tt$conf.int[1], upper = tt$conf.int[2])
```

Full-time nurses earn `r sprintf("%.2f", mean(matched$hourly_rate_left - matched$hourly_rate_right))`
more per hour than their matched part-time counterparts, with a 95 percent
interval that excludes zero. As a paired difference on a matched sample this
estimates an adjusted association. Reading it as the causal effect of full-time
status is what rests on assumptions no matching method can verify: conditional
ignorability, positivity, and the stable unit treatment value assumption
(SUTVA), under which a unit's outcome depends on its own treatment alone. The
matched output can also be passed to an outcome model for regression
adjustment; valid estimation and uncertainty quantification still depend on the
estimator used.

What matching can offer against the first of those assumptions is a statement
about how strong an unmeasured confounder would have to be to overturn the
result. `sensitivity_analysis()` implements Rosenbaum bounds (Rosenbaum 2002)
for the Wilcoxon signed-rank statistic on one-to-one matched pairs, reporting
upper and lower p-value bounds across a grid of the odds-ratio parameter
$\Gamma$.

```{r estimate-sensitivity, echo=TRUE}
sens <- sensitivity_analysis(m, treated, control, outcome_var = "hourly_rate")
sens$results[sens$results$gamma %in% c(1, 1.5, 1.75, 2), ]
```

At $\Gamma = 1$, meaning no hidden bias, the association is significant. It
survives an unmeasured confounder that raises one pair member's odds of being
full-time by half, and stops being significant at
$\Gamma =$ `r sens$critical_gamma`. A reader can judge from that number whether
a confounder of that strength is plausible in this setting; the analysis
quantifies exposure to hidden bias rather than removing it.

\FloatBarrier

# Solver selection across problem regimes

`assignment(method = "auto")` chooses a solver from five rules read off one pass
over the cost matrix: enumerate an at most 8 by 8 problem, use a cardinality
algorithm where the finite costs are constant or binary, use the sparse
Jonker-Volgenant variant where more than half the entries are forbidden, use the
shortest-augmenting-path solver where there are at least three times as many
columns as rows, and otherwise use Jonker-Volgenant. The article's solver figure
varies problem size on square dense integer costs, which is the last rule's
regime and no other. This grid varies the properties the rules are stated over.

A cell is a cost regime, a forbidden pattern and a shape. Each cell is generated
as several independent instances, each instance is timed several times, and
every solver in the panel sees the same instance.

```{r regime-design}
reg_lab <- c(int_uniform = "integer, uniform on 1..10,000",
             dbl_uniform = "double, uniform on (0, 1)",
             binary = "binary, 0 or 1",
             constant = "constant",
             tied = "integer, five distinct values",
             heavy_tailed = "double, lognormal",
             metric_uniform = "euclidean over uniform points in 8 dimensions",
             metric_clustered = "euclidean over eight clusters in 8 dimensions")
pat_lab <- c(none = "complete: every pair admissible",
             random_60 = "60 percent of pairs admissible, unstructured",
             random_25 = "25 percent admissible, unstructured",
             block_4 = "25 percent admissible in four disconnected components")

kb(data.frame(Factor = c(rep("cost regime", length(reg_lab)),
                         rep("forbidden pattern", length(pat_lab))),
              Level = c(names(reg_lab), names(pat_lab)),
              Meaning = c(unname(reg_lab), unname(pat_lab))),
   caption = "The factors the solver grid crosses. Shapes and sizes are in the next table.",
   align = "lll")
```

```{r regime-shapes}
sh <- unique(regime_cells[, c("tier", "n_rows", "n_cols")])
sh <- sh[order(sh$tier, sh$n_rows, sh$n_cols), ]
sh$aspect <- sprintf("1:%.0f", sh$n_cols / sh$n_rows)
sh$Shape <- shape(sh$n_rows, sh$n_cols)
tier_panel <- tapply(regime_runs$method, regime_runs$tier,
                     function(x) length(unique(x)))
sh$`Solvers in the panel` <- as.integer(tier_panel[sh$tier])
kb(sh[, c("tier", "Shape", "aspect", "Solvers in the panel")],
   caption = "Shapes. The base tier runs the whole panel at a size every solver in it reaches; the large tier repeats the crossing at a size only the scalable solvers reach.",
   align = "llrr")
```

## What the rules cost

For each cell, the solver `"auto"` selected is timed as a solver of its own, so
its number carries the probing pass the rules cost, and it is compared against
the fastest named solver in the same cell.

```{r regime-by-rule}
by_rule <- do.call(rbind, lapply(split(regime_verdict, regime_verdict$auto_rule),
                                 function(g) data.frame(
  Rule = g$auto_rule[1],
  Cells = nrow(g),
  `Picked the cell's fastest` = sum(g$auto_picks == g$best_method),
  `Median ratio` = sprintf("%.2fx", median(g$ratio)),
  `Worst ratio` = sprintf("%.2fx", max(g$ratio)),
  check.names = FALSE)))
kb(by_rule[order(-by_rule$Cells), ],
   caption = "Each dispatch rule, over the cells where it fired. Ratio is the time the dispatched solver took over the time the cell's fastest named solver took, so 1.00x means the rule picked the best available solver and 2.00x means it cost twice the best.",
   align = "lrrrr")
```

```{r regime-worst}
w <- regime_verdict[order(-regime_verdict$ratio), ]
w <- head(w, 25)
kb(data.frame(Regime = w$regime, Pattern = short_pat(w$pattern),
              Shape = gsub(" ", "", w$shape), Rule = w$auto_rule,
              Picked = w$auto_picks,
              `Best` = w$best_method,
              `Dispatched (ms)` = sprintf("%.3g", 1000 * w$auto_s),
              `Best (ms)` = sprintf("%.3g", 1000 * w$best_s),
              Ratio = sprintf("%.2fx", w$ratio), check.names = FALSE),
   caption = "The 25 cells where the dispatched solver was furthest off the cell's fastest.",
   align = "llllllrrr")
```

## Every cell

```{r regime-all}
v <- regime_verdict[order(regime_verdict$tier, regime_verdict$regime,
                          regime_verdict$pattern, regime_verdict$shape), ]
kb(data.frame(Regime = v$regime, Pattern = short_pat(v$pattern),
              Shape = gsub(" ", "", v$shape), Rule = v$auto_rule,
              Picked = v$auto_picks, Best = v$best_method,
              Ratio = sprintf("%.2fx", v$ratio), check.names = FALSE),
   caption = "Every cell of the solver grid: the rule that fired, the solver it picked, the fastest solver in the cell, and the ratio between them.",
   align = "llllllr")
```

## Every solver, every cell

```{r regime-full}
r <- regime_results[order(regime_results$tier, regime_results$regime,
                          regime_results$pattern, regime_results$n_rows,
                          regime_results$n_cols, regime_results$median_s), ]
kb(data.frame(Regime = r$regime, Pattern = short_pat(r$pattern),
              Shape = shape(r$n_rows, r$n_cols), Solver = r$method,
              Instances = r$instances,
              `Median (ms)` = sprintf("%.3g", 1000 * r$median_s),
              `Q25 (ms)` = sprintf("%.3g", 1000 * r$q25_s),
              `Q75 (ms)` = sprintf("%.3g", 1000 * r$q75_s),
              check.names = FALSE),
   caption = "Milliseconds per solver and cell: the median across instances, with the quartiles beside it. An instance contributes the median of its own repetitions, so the quartiles are between problems rather than between repetitions of one problem. A solver that errored or timed out on a cell has no row there.",
   align = "llllrrrr")
```

## Do the solvers agree

Every solver in a cell solves the same instance, so their totals have to agree.
A disagreement would be a defect in a solver rather than a property of the
regime.

```{r regime-agreement}
ok <- regime_runs[regime_runs$status == "ok" & !is.na(regime_runs$total_cost), ]
key <- paste(ok$tier, ok$regime, ok$pattern, ok$n_rows, ok$n_cols, ok$instance)
ag <- do.call(rbind, lapply(split(ok, key), function(g) {
  ref <- median(g$total_cost)
  data.frame(worst = max(abs(g$total_cost - ref) / pmax(1, abs(ref))),
             solvers = length(unique(g$method)))
}))
cat(sprintf("%d instances carrying %d to %d solvers each; the worst relative disagreement on the optimum is %.3g.\n",
            nrow(ag), min(ag$solvers), max(ag$solvers), max(ag$worst)))
```

\FloatBarrier

# The edge-generation loop away from one cloud

The article's implicit results come from one cloud: eight independent normal
covariates with the treated group shifted, one draw per size. This grid sweeps
one factor at a time around that base, several seeds deep. Where the dense path
can afford the problem, the loop's answer is compared against it unit by unit.

```{r igrid-summary}
g <- igrid[order(igrid$sweep, igrid$n_total, igrid$dim, igrid$cloud,
                 igrid$caliper), ]
kb(data.frame(Sweep = g$sweep, Cloud = g$cloud, Dim = g$dim,
              Units = format(g$n_total, big.mark = ","), Caliper = g$caliper,
              Seeds = g$seeds,
              Rounds = sprintf("%g-%g", g$rounds_min, g$rounds_max),
              `Graph %` = sprintf("%.3f", g$graph_pct_med),
              `Distances` = sprintf("%.2fx", g$distances_x_med),
              `Implicit` = sprintf("%.3g s", g$implicit_s_med),
              `Lazy` = sprintf("%.3g s", g$lazy_s_med),
              `Speedup` = sprintf("%.2fx", g$speedup_med),
              check.names = FALSE),
   caption = "The loop by cell. Rounds is the range over seeds. Graph is the median share of the complete pair count the candidate set ended up holding, in percent. Distances is the median number of distance evaluations over all rounds as a multiple of the complete pair count. Implicit and lazy are median seconds over seeds and speedup is their median ratio.",
   align = "lllllrrrrrrr")
```

```{r igrid-certificate}
kb(data.frame(Sweep = g$sweep, Cloud = g$cloud, Dim = g$dim,
              Units = format(g$n_total, big.mark = ","), Caliper = g$caliper,
              `Worst duality gap` = format(g$worst_gap, digits = 3),
              `All certified` = g$all_certified,
              `Equals the dense solve` = ifelse(is.na(g$equal_to_dense),
                                                "not run", g$equal_to_dense),
              check.names = FALSE),
   caption = "The certificate by cell. The duality gap is the residual the loop stopped on. Equals the dense solve compares the pairing unit by unit where the dense path could be run, and reads not run where the problem was too large for it.",
   align = "lllllrrr")
```

```{r igrid-spread}
imp_runs <- igrid_runs[igrid_runs$memory_mode == "implicit" &
                         igrid_runs$status == "ok", ]
imp_runs$graph_pct <- 100 * imp_runs$candidate_edges / imp_runs$possible_edges
kb(data.frame(Sweep = imp_runs$sweep, Cloud = imp_runs$cloud,
              Dim = imp_runs$dim,
              Units = format(imp_runs$n_total, big.mark = ","),
              Caliper = imp_runs$caliper, Seed = imp_runs$seed_index,
              Rounds = imp_runs$n_rounds,
              `Seed width` = imp_runs$seed_width,
              `Graph %` = sprintf("%.3f", imp_runs$graph_pct),
              `Seconds` = sprintf("%.3g", imp_runs$elapsed_s),
              `Gap` = format(imp_runs$duality_gap, digits = 3),
              check.names = FALSE),
   caption = "Every implicit run in the grid, one row per seed. Seed width is the number of columns the first round gave each row.",
   align = "lllllrrrrrr")
```

```{r igrid-failures}
bad <- igrid_runs[igrid_runs$status != "ok", ]
if (nrow(bad)) {
  kb(data.frame(Sweep = bad$sweep, Cloud = bad$cloud, Dim = bad$dim,
                Units = bad$n_total, Caliper = bad$caliper,
                Mode = bad$memory_mode, Seed = bad$seed_index,
                Status = bad$status, check.names = FALSE),
     caption = "Cells that did not finish.", align = "llllllll")
} else {
  cat("Every cell of the grid finished.\n")
}
```

\FloatBarrier

# Peak memory

Peak resident set size is a property of a process, so each cell here is a fresh
R process that loads what it needs, does one matching and exits, measured from
outside by the platform's `time` utility. The baseline row of a size is a
process that loaded the same packages and matched nothing, and it is what the
increment column is measured against. `dense_matrix` builds the distance matrix
and stops, which separates the representation from the solver that runs on it.

```{r memory-table}
m <- memory[order(memory$n_total, memory$arm), ]
kb(data.frame(Units = format(m$n_total, big.mark = ","), Arm = m$arm,
              `Peak RSS` = sprintf("%.0f MB", m$peak_rss_mb),
              `Over baseline` = ifelse(is.na(m$over_baseline_mb), "-",
                                       sprintf("%.0f MB", m$over_baseline_mb)),
              `R heap peak` = sprintf("%.0f MB", m$heap_peak_mb),
              `Matrix` = sprintf("%.0f MB", m$matrix_mb),
              `Package estimate` = sprintf("%.0f MB", m$estimate_mb),
              `Seconds` = sprintf("%.3g", m$elapsed_s),
              Status = m$status, check.names = FALSE),
   caption = "Peak resident set size by arm and problem size. Matrix is the bytes the dense cost matrix itself needs at that shape. Package estimate is what estimate_dense_matrix_mb() predicts, which is four times the matrix and is what the memory dispatcher warns from.",
   align = "llrrrrrrl")
```

```{r memory-heuristic}
dm <- memory[memory$arm == "dense_matrix" & !is.na(memory$over_baseline_mb), ]
if (nrow(dm)) {
  kb(data.frame(Units = format(dm$n_total, big.mark = ","),
                `Matrix` = sprintf("%.0f MB", dm$matrix_mb),
                `Measured over baseline` = sprintf("%.0f MB", dm$over_baseline_mb),
                `Measured / matrix` = sprintf("%.2f", dm$over_baseline_mb / dm$matrix_mb),
                `Package estimate` = sprintf("%.0f MB", dm$estimate_mb),
                check.names = FALSE),
     caption = "The four-fold estimate against the measurement. The estimate is deliberately conservative: it exists to avoid under-warning before a large allocation, not to predict the peak.",
     align = "lrrrr")
}
```

\FloatBarrier

# Representation against solver

The implicit mode is a representation and a solver at once: its restricted
master is the flow model, and naming a solver under it is an error rather than a
setting. Reading its timing as a representation comparison would therefore
attribute the flow solver's behaviour to the representation. The table below is
the decomposition: down a column, one solver over two representations; across a
row, one representation under two solvers.

```{r repr-solver}
sizes <- sort(unique(imp$n_total))
cell <- function(mode, n) {
  v <- imp$elapsed_s[imp$memory_mode == mode & imp$n_total == n]
  if (!length(v) || is.na(v[1])) return("not run")
  if (v[1] < 1) sprintf("%.0f ms", 1000 * v[1]) else sprintf("%.3g s", v[1])
}
kb(data.frame(
  Units = format(sizes, big.mark = ","),
  `Dense, JV` = vapply(sizes, function(n) cell("dense", n), character(1)),
  `Lazy, JV` = vapply(sizes, function(n) cell("lazy", n), character(1)),
  `Dense, flow` = vapply(sizes, function(n) cell("dense_csflow", n), character(1)),
  `Implicit` = vapply(sizes, function(n) cell("implicit", n), character(1)),
  check.names = FALSE),
  caption = "Wall-clock seconds by representation and solver. Dense and lazy differ only in whether the matrix is materialised; dense and dense-flow differ only in the solver; implicit is the mode as a user meets it.",
  align = "lrrrr")
```

\FloatBarrier

# Scaling, every run

The scaling table in the article reports the median across instances. These are
the runs behind it, one row per repetition.

```{r scaling-runs}
sr <- scaling_runs[order(scaling_runs$n_total, scaling_runs$package,
                         scaling_runs$instance, scaling_runs$rep), ]
kb(data.frame(Units = format(sr$n_total, big.mark = ","), Package = sr$package,
              Instance = sr$instance, Seed = sr$seed, Rep = sr$rep,
              Seconds = ifelse(is.na(sr$seconds), "-",
                               sprintf("%.4g", sr$seconds)),
              Status = sr$status, check.names = FALSE),
   caption = "Every scaling run. Instance is an independently generated problem of that size; rep is a repetition on that problem. A run that failed or timed out carries its status and no time.",
   align = "llrrrrl")
```

\FloatBarrier

# Software design

The article describes the three layers the package is organised in, the dispatch
rules, the memory modes and how the package is verified. This section carries the
object model, the C++ organisation and the dependency footprint.

## Object model

The object model is S3 throughout. Matching results carry the class vector
`c("matching_result", "couplr_result")`, with the design-specific classes
(`full_matching_result`, `cem_result`, `subclass_result`) sharing the
`couplr_result` parent, which is what lets `balance_diagnostics()`,
`join_matched()`, `match_data()` and the `bal.tab()` methods dispatch on it once
rather than being rewritten per design. Results are plain lists of tibbles, so a
user who wants something the package does not provide can reach in and take it.
`lap_solve()` returns a tibble subclassed as `lap_solve_result` with `source`,
`target` and `cost` columns, so an optimum is usable in a plot or a join without
an extraction step, while `get_total_cost()` and `get_method_used()` recover the
metadata carried in attributes.

## C++ organisation

All nineteen solvers are written from scratch in C++ and exposed through Rcpp,
with `src/core` holding the cost-source types and shared utilities with no Rcpp
dependency, `src/solvers` one algorithm per file with a thin `_rcpp.cpp` wrapper
each, `src/flow` the flow model, its two solvers and the edge-generation loop,
and `src/interface` the conversion between R objects and the internal types. The
separation means the solver bodies are ordinary C++ that can be compiled and
tested outside R, which is what the C++ test suite does.

Solvers are templated on their cost source rather than written against a concrete
matrix. A cost source provides `at(i, j)`, `allowed(i, j)` and `empty()`, and two
types satisfy that interface: `CostMatrix`, which owns a dense flat array plus a
mask, and `LazyCostMatrix`, which owns the two feature matrices and computes
$C_{ij}$ on demand from the requested metric, applying calipers and the distance
ceiling in `allowed()`. A handful of hot loops index the dense mask array
directly for speed, which the lazy type cannot support; these are guarded by a
`supports_raw_mask` trait and an `if constexpr` branch rather than by duplicating
the solver. One solver body therefore serves both cost representations, and the
pricing oracle of the edge-generation loop needs `at(i, j)` and `allowed(i, j)`
and nothing else, so it consumes the same concept and adds no third
representation of the costs.

`LazyCostMatrix` also bakes in the transformations that the dense path applies in
a separate preparation pass, namely mapping forbidden entries to a large sentinel
and negating costs when maximising. Doing this at construction is what allows the
untouched solver body to work on either type.

## Dependency footprint

couplr declares Rcpp, tibble, dplyr, rlang, purrr and `methods` in `Imports`, and
`Rcpp` alone in `LinkingTo`. The recursive closure of hard dependencies is 17
non-base packages, so a default installation brings 18 packages including couplr
itself. The comparable figures on the same day were 18 for optmatch, 8 for
MatchIt and 7 for designmatch.

Two choices sit behind that number. First, we use no external solver: all
nineteen algorithms are package source, so there is no dependency on an LP or MIP
library and no system solver to install. That choice also keeps `LinkingTo` to
`Rcpp` alone, which matters because every entry there forces a compile of that
package before couplr can be built from source; couplr declares only what its
headers actually include. Second, everything that serves a single feature lives
in `Suggests` and is loaded at call time. The animation and image-morphing
functions, the plotting helpers, the interoperability layer for MatchIt, optmatch
and cobalt, and the parallel back end are all optional in this sense, so a user
who only wants to match pays for none of them.
