---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(couplr)
library(dplyr)
```

## What couplr Does

couplr creates matched pairs from two groups of observations. Given a "left" group (e.g., treatment) and a "right" group (e.g., control), it finds the best one-to-one pairing based on similarity across variables you specify.

**Common use cases:**

- Matching treated patients to similar controls in observational studies
- Pairing survey respondents for comparison
- Creating balanced samples for causal inference

### Documentation Roadmap

| Vignette | Focus | Audience |
|----------|-------|----------|
| **Quick Start** (this) | Basic matching with `match_couples()` | Everyone |
| [Matching Workflows](matching-workflows.html) | Full pipeline: preprocessing, blocking, diagnostics | Researchers |
| [Algorithms](algorithms.html) | Mathematical foundations, solver selection | Technical users |
| [Comparison](comparison.html) | vs MatchIt, optmatch, designmatch | Package evaluators |

**Start here**, then proceed to whichever vignette matches your use case.

---

## Your First Match

The simplest workflow uses `match_couples()`:
```{r first-match}
library(couplr)
library(dplyr)

# Create example data: treatment and control groups
set.seed(123)
treatment <- tibble(
  id = 1:50,
  age = rnorm(50, mean = 45, sd = 10),
  income = rnorm(50, mean = 55000, sd = 12000)
)

control <- tibble(
  id = 1:80,
  age = rnorm(80, mean = 50, sd = 12),
  income = rnorm(80, mean = 48000, sd = 15000)
)

# Match on age and income
result <- match_couples(
  left = treatment,
  right = control,
  vars = c("age", "income"),
  auto_scale = TRUE
)

# View matched pairs
head(result$pairs)
```

**What happened:**

1. couplr calculated how similar each treatment unit is to each control unit
2. It found the optimal one-to-one pairing that minimizes total distance
3. Each treatment unit gets matched to exactly one control unit

### Understanding the Output

```{r output-explained}
# Quick overview with summary()
summary(result)

# Or access specific info
result$info$n_matched
```

The `result$pairs` table contains:

- `left_id`: Row number from the treatment group
- `right_id`: Row number from the control group
- `distance`: How different the matched units are (lower = more similar)

---

## Why Scaling Matters

Without scaling, variables with larger values dominate the matching. Income (measured in thousands) would overwhelm age (measured in decades):

```{r scaling-demo}
# BAD: Without scaling, income dominates
result_unscaled <- match_couples(
  treatment, control,
  vars = c("age", "income"),
  auto_scale = FALSE
)

# GOOD: With scaling, both variables contribute equally
result_scaled <- match_couples(
  treatment, control,
  vars = c("age", "income"),
  auto_scale = TRUE
)

# Compare mean distances
cat("Unscaled mean distance:", round(mean(result_unscaled$pairs$distance), 1), "\n")
cat("Scaled mean distance:", round(mean(result_scaled$pairs$distance), 3), "\n")
```

**Rule of thumb:** Always use `auto_scale = TRUE` unless you have a specific reason not to.

---

## Checking Match Quality

After matching, verify that treatment and control groups are now balanced:

```{r balance-check}
# Get the matched observations
matched_treatment <- treatment[result$pairs$left_id, ]
matched_control <- control[result$pairs$right_id, ]

# Compare means before and after matching
cat("BEFORE matching:\n")
cat("  Age difference:", round(mean(treatment$age) - mean(control$age), 1), "years\n")
cat("  Income difference: $", round(mean(treatment$income) - mean(control$income), 0), "\n\n")

cat("AFTER matching:\n")
cat("  Age difference:", round(mean(matched_treatment$age) - mean(matched_control$age), 1), "years\n")
cat("  Income difference: $", round(mean(matched_treatment$income) - mean(matched_control$income), 0), "\n")
```

For formal balance assessment, use `balance_diagnostics()` (covered in [Matching Workflows](matching-workflows.html)).

### Visualizing Match Quality

Use `plot()` to see the distribution of match distances:

```{r plot-result, fig.width=6, fig.height=4}
plot(result)
```

The histogram shows how similar matched pairs are. A distribution concentrated near zero indicates good matches.

---

## Large Datasets: Use Greedy Matching

For datasets larger than a few thousand observations, optimal matching becomes slow. Use `greedy_couples()` instead---it's 10-100x faster with nearly identical results:

```{r greedy-example}
# Create larger datasets
set.seed(456)
large_treatment <- tibble(
  id = 1:2000,
  age = rnorm(2000, 45, 10),
  income = rnorm(2000, 55000, 12000)
)

large_control <- tibble(
  id = 1:3000,
  age = rnorm(3000, 50, 12),
  income = rnorm(3000, 48000, 15000)
)

# Fast greedy matching
result_greedy <- greedy_couples(
  large_treatment, large_control,
  vars = c("age", "income"),
  auto_scale = TRUE,
  strategy = "row_best"  # fastest strategy
)

cat("Matched", result_greedy$info$n_matched, "pairs\n")
cat("Mean distance:", round(mean(result_greedy$pairs$distance), 3), "\n")
```

**When to use which:**

| Dataset size | Recommended function |
|--------------|---------------------|
| < 1,000 per group | `match_couples()` |
| 1,000 - 5,000 | Either works; greedy is faster |
| > 5,000 | `greedy_couples()` |

---

## Setting a Maximum Distance (Caliper)

Sometimes you want to reject poor matches rather than force bad pairings. Use `max_distance` to set a caliper:

```{r caliper-example}
# Allow any match
result_loose <- match_couples(
  treatment, control,
  vars = c("age", "income"),
  auto_scale = TRUE
)

# Only allow close matches
result_strict <- match_couples(
  treatment, control,
  vars = c("age", "income"),
  auto_scale = TRUE,
  max_distance = 0.5  # reject pairs more different than this
)

cat("Without caliper:", result_loose$info$n_matched, "pairs\n")
cat("With caliper:", result_strict$info$n_matched, "pairs\n")
```

Stricter calipers mean fewer but better matches.

---

## Complete Example

Here's a realistic workflow from start to finish:

```{r complete-example}
# 1. Prepare your data
set.seed(789)
patients_treated <- tibble(
  patient_id = paste0("T", 1:100),
  age = rnorm(100, 62, 8),
  bmi = rnorm(100, 28, 4),
  smoker = sample(0:1, 100, replace = TRUE, prob = c(0.6, 0.4))
)

patients_control <- tibble(
  patient_id = paste0("C", 1:200),
  age = rnorm(200, 58, 10),
  bmi = rnorm(200, 26, 5),
  smoker = sample(0:1, 200, replace = TRUE, prob = c(0.7, 0.3))
)

# 2. Match on clinical variables
matched <- match_couples(
  left = patients_treated,
  right = patients_control,
  vars = c("age", "bmi", "smoker"),
  auto_scale = TRUE,
  max_distance = 1.0  # optional caliper
)

# 3. Check how many matched
cat("Treated patients:", nrow(patients_treated), "\n")
cat("Successfully matched:", matched$info$n_matched, "\n")
cat("Match rate:", round(100 * matched$info$n_matched / nrow(patients_treated), 1), "%\n")

# 4. Extract matched samples for analysis
treated_matched <- patients_treated[matched$pairs$left_id, ]
control_matched <- patients_control[matched$pairs$right_id, ]

# 5. Verify balance
cat("\nBalance check (difference in means):\n")
cat("  Age:", round(mean(treated_matched$age) - mean(control_matched$age), 2), "\n")
cat("  BMI:", round(mean(treated_matched$bmi) - mean(control_matched$bmi), 2), "\n")
cat("  Smoker %:", round(100*(mean(treated_matched$smoker) - mean(control_matched$smoker)), 1), "\n")
```

---

## Next Steps

You now know the basics of matching with couplr. Here's where to go next:

**For production research workflows:**

- [Matching Workflows](matching-workflows.html) covers preprocessing, blocking, formal balance diagnostics, and publication-ready output

**For understanding algorithm choices:**

- [Algorithms](algorithms.html) explains when different solvers are faster or more appropriate

**For comparing with other packages:**

- [Comparison](comparison.html) shows how couplr differs from MatchIt, optmatch, and designmatch

---

## Additional: Direct Assignment Problem Solving

If you need to solve assignment problems directly (not matching workflows), couplr also provides lower-level functions.

### lap_solve(): Matrix-Based Assignment

Given a cost matrix where entry (i,j) is the cost of assigning row i to column j:

```{r lap-solve-basic}
# Cost matrix: 3 workers × 3 tasks
cost <- matrix(c(
  4, 2, 5,
  3, 3, 6,
  7, 5, 4
), nrow = 3, byrow = TRUE)

result <- lap_solve(cost)
print(result)
```

Row 1 is assigned to column 2 (cost 2), row 2 to column 1 (cost 3), row 3 to column 3 (cost 4). Total cost: 9.

### Forbidden Assignments

Use `NA` or `Inf` for impossible assignments:

```{r forbidden}
cost_forbidden <- matrix(c(
  4, 2, NA,   # Row 1 cannot go to column 3
  Inf, 3, 6,  # Row 2 cannot go to column 1
  7, 5, 4
), nrow = 3, byrow = TRUE)

lap_solve(cost_forbidden)
```

### Maximization

For preference or profit maximization:

```{r maximize}
preferences <- matrix(c(
  8, 5, 3,
  4, 7, 6,
  2, 4, 9
), nrow = 3, byrow = TRUE)

lap_solve(preferences, maximize = TRUE)
```

### Batch Solving

Solve many independent problems at once:

```{r batch}
# 50 random cost matrices
cost_list <- lapply(1:50, function(i) matrix(runif(9, 1, 10), 3, 3))

batch_results <- lap_solve_batch(cost_list)

# Summary
batch_results |>
  distinct(problem_id, total_cost) |>
  summarise(
    problems = n(),
    mean_cost = mean(total_cost)
  )
```

### K-Best Solutions

Find multiple near-optimal solutions:

```{r kbest}
cost <- matrix(c(1, 2, 3, 4, 3, 2, 5, 4, 1), nrow = 3, byrow = TRUE)

kbest <- lap_solve_kbest(cost, k = 3)
print(kbest)
```

---

## See Also

- `?match_couples` - Optimal matching function reference
- `?greedy_couples` - Fast approximate matching
- `?balance_diagnostics` - Formal balance assessment
- `?lap_solve` - Direct assignment problem solving
