---
title: "couplr: Optimal Pairing and Matching via Linear Assignment"
date: "`r Sys.Date()`"
abstract: >
  Matching two groups of units so that the total cost of the chosen pairs is as
  small as possible is the linear assignment problem, which sits underneath
  causal-inference matching, experimental design, task allocation and image
  alignment. R users currently meet it through two separate families of
  packages: causal-inference packages that reach the solver through a
  design-level argument, and solver packages that expose one algorithm each
  without preprocessing, constraints or balance diagnostics. We present the
  couplr package, which covers both. A data-frame interface builds distances,
  applies calipers, blocks and forbidden pairs, and returns matched output with
  balance diagnostics. A matrix interface exposes nineteen assignment
  algorithms written from scratch in C++, together with dual variables that
  certify optimality, k-best assignments and bottleneck assignment. A
  dispatcher chooses among the solvers from problem shape, sparsity, cost type
  and size, and a lazy cost representation matches fifty thousand units without
  building the distance matrix. Constrained problems are solved to their
  lexicographic optimum, the largest admissible matching and the cheapest among
  those of its size. On a one-to-one Mahalanobis task couplr reproduces the
  balance MatchIt and optmatch achieve, at sizes where both stop. couplr is
  available on CRAN.
draft: false
author:
  - name: Gilles Colling
    affiliation: University of Vienna
    address:
    - Department of Botany and Biodiversity Research
    - Rennweg 14, 1030 Vienna, Austria
    url: https://gillescolling.com
    orcid: 0000-0003-3070-6066
    email: gilles.colling051@gmail.com
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
header-includes:
  - \usepackage{placeins}
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)
library(couplr)
library(ggplot2)

bench     <- read.csv("data/benchmark-table.csv", stringsAsFactors = FALSE)
lalonde   <- read.csv("data/lalonde-per-covariate.csv", stringsAsFactors = FALSE)
lal_obj   <- read.csv("data/lalonde-results.csv", stringsAsFactors = FALSE)
scaling   <- read.csv("data/scaling-results.csv", stringsAsFactors = FALSE)
lazy      <- read.csv("data/scaling-lazy-results.csv", stringsAsFactors = FALSE)
couplr_ver <- as.character(utils::packageVersion("couplr"))
```

# Introduction

Many data-analysis tasks come down to the same question: given two sets of
objects and a cost attached to every possible pair, which pairing has the
smallest total cost? A researcher building an observational study pairs treated
units with controls that resemble them on measured covariates. A scheduler
assigns staff to shifts, an ecologist pairs surveyed plots across two time
points, and an image-processing pipeline maps the pixels of one picture onto
the pixels of another. In each case the mathematical object is the linear
assignment problem, and the quality of the answer depends on solving it to
optimality rather than greedily.

R has good software at both ends of this problem, in two separate groups of
packages. On the statistical side, \CRANpkg{MatchIt} [@MatchIt] provides a
common interface to several matching designs and delegates optimal pair and
full matching to \CRANpkg{optmatch} [@optmatch]. \pkg{optmatch} formulates
those designs as minimum-cost flow problems [@HansenKlopfer2006] and supports
more than one back end, currently RELAX-IV and LEMON, selected through its
`solver` argument. \CRANpkg{Matching} [@Matching] provides multivariate and
propensity-score matching, with `GenMatch()` using genetic search to optimise
covariate weights. \CRANpkg{designmatch} [@designmatch] formulates
balance-constrained designs as integer programs and defaults to the open-source
HiGHS solver. \CRANpkg{cobalt} [@cobalt] supplies balance diagnostics across
packages. What these expose is the matching design; the assignment problem
underneath is reached through a solver argument at most, without dual
variables, alternative optima or a bottleneck objective. Statements about other
packages here and in Table \@ref(tab:capability) were assessed on 2026-08-09
against \pkg{MatchIt} 4.7.2, \pkg{optmatch} 0.10.8, \pkg{Matching} 4.10-15 and
\pkg{designmatch} 0.5.5.

On the algorithmic side, \CRANpkg{clue} [@clue] exposes the Hungarian method
through `solve_LSAP()`, and \CRANpkg{lpSolve} [@lpSolve] solves the assignment
problem as a general linear program. These return an optimum for a cost matrix
the user has already built; covariate handling, calipers and balance tables sit
outside their scope.

A workflow that needs both halves therefore threads several packages by hand.
Matching on observed covariates instead of a propensity score, imposing
calipers on more than one variable at a time, blocking, and choosing the solver
deliberately are all available in R, each in a different package.

We present the \CRANpkg{couplr} package [@couplr], which treats the assignment
problem itself as the organising object and builds the workflow around it. A
user who starts from a data frame with covariates and a treatment indicator,
and a user who starts from a cost matrix, enter through different functions and
reach the same solver core. We contribute three things to R:

1. A matching interface in which covariate distance is the primary entry point,
   with calipers, blocking, forbidden pairs, variable ratios, replacement,
   balance diagnostics and Rosenbaum sensitivity bounds in one object model.
2. Nineteen assignment algorithms implemented from scratch in C++ and reachable
   by name, covering augmenting-path, auction, cost-scaling and network-flow
   families, together with dual variables, k-best assignments, bottleneck
   assignment and entropy-regularised transport.
3. A dispatcher and a memory model that pick the solver and the cost
   representation from the problem, so the interface does not change as
   problems grow from hundreds to tens of thousands of units.

We state the problem couplr solves and the algorithm families that solve it,
walk through a matching workflow and the designs around it, then describe the
solver layer. The software design follows in detail: object model, C++
organisation, dispatch rules, memory modes, verification and dependency
footprint. A performance section reports per-solver timings, a balance
comparison against \pkg{MatchIt} and \pkg{optmatch} on the LaLonde data, and a
scaling comparison. Results were produced with couplr version `r couplr_ver`.

# The assignment problem behind matching

Let $C \in \mathbb{R}^{n \times m}$ be a cost matrix whose entry $C_{ij}$ is the
cost of pairing row $i \in \{1, \dots, n\}$ with column $j \in \{1, \dots, m\}$,
and take $n \le m$; couplr transposes the problem internally when it is not and
maps the assignment back afterwards, so the user never sees the transpose.
Write $X \in \{0,1\}^{n \times m}$ for the assignment matrix, with $X_{ij} = 1$
when row $i$ is matched to column $j$. The linear assignment problem is

\begin{equation}
\min_{X} \; \sum_{i=1}^{n} \sum_{j=1}^{m} C_{ij} X_{ij}
\quad \text{subject to} \quad
\sum_{j} X_{ij} = 1 \; \forall i, \qquad
\sum_{i} X_{ij} \le 1 \; \forall j .
(\#eq:lap)
\end{equation}

Every row is matched and every column is used at most once. The row constraint
is an equality: relaxing it to $\sum_j X_{ij} \le 1$ would leave $X = 0$
optimal for any non-negative cost matrix, which is not the problem a matching
application poses.

In a matching application, rows are treated units, columns are controls, and
$C_{ij}$ is a distance between covariate vectors. In a scheduling application,
rows are workers, columns are shifts, and $C_{ij}$ is a cost or the negative of
a preference score. Forbidden pairs are encoded as $C_{ij} = \infty$, and
enough of them make \@ref(eq:lap) infeasible, since no assignment can then
reach every row. couplr solves a lexicographic relaxation in that case: the
largest number of matched rows first, then the smallest total cost among the
assignments attaining that number. The section on constraints below describes
how that problem is reached and what it costs.

The constraint matrix of \@ref(eq:lap) is totally unimodular, so the linear
programming relaxation has integral optima [@Birkhoff1946; @Burkard1999] and the
combinatorial problem can be solved by primal-dual methods rather than by branch
and bound. The dual assigns a potential $u_i$ to each row and $v_j$ to each
column subject to

\begin{equation}
u_i + v_j \le C_{ij} \quad \forall (i,j), \qquad
u_i + v_j = C_{ij} \quad \text{whenever } X_{ij} = 1 ,
(\#eq:duals)
\end{equation}

Dual feasibility on its own says nothing about a particular $X$. The certificate
needs four parts together: $X$ is primal feasible, $(u, v)$ satisfies the
inequality in \@ref(eq:duals), every selected edge is tight, and the primal and
dual objectives agree. couplr returns the potentials from `assignment_duals()`,
so a user can check all four without trusting the solver. The reduced cost
$C_{ij} - u_i - v_j$ is a lower bound on what forcing pair $(i,j)$ would add to
the total rather than the whole of it, because forcing one pair can displace
others.

## Why more than one algorithm

Every solver of \@ref(eq:lap) returns the same optimal value, so the choice
among them is a performance question rather than a statistical one, and the
performance depends on the shape of the problem. We group the solvers into four
families and a set of special-purpose routines.

**Augmenting-path methods** grow a matching one row at a time, each time finding
a shortest augmenting path in the residual graph under reduced costs. The
Hungarian method [@Kuhn1955; @Munkres1957] is the classical member; the
Jonker-Volgenant algorithm [@JonkerVolgenant1987] adds column reduction and
auction-style initialisation, which is what makes it fast in practice on dense
problems despite an $O(n^3)$ worst case. `lapmod` is the sparse variant, working
from an adjacency list so that forbidden pairs cost nothing to carry.

**Auction methods** [@Bertsekas1988] let rows bid for columns, raising prices
until no row wants to switch. They converge to optimality once the bid increment
$\varepsilon$ falls below $1/n$ for integer costs, and $\varepsilon$-scaling
gives the useful practical variants. Auction is naturally parallel and tolerant
of near-degenerate cost structures where augmenting-path methods spend time on
long paths.

**Cost-scaling methods** solve a sequence of increasingly precise problems,
each time halving the scale on which costs are rounded and reusing the previous
solution [@GoldbergKennedy1995]. The Gabow-Tarjan bit-scaling algorithm
[@GabowTarjan1989] belongs here, and reaches
$O(\sqrt{n}\, m \log(n C_{\max}))$, where $C_{\max} = \max_{i,j} |C_{ij}|$.
To our knowledge this algorithm did not previously have a publicly available
open-source implementation in any language.

**Network-flow methods** treat \@ref(eq:lap) as a minimum-cost flow problem on a
bipartite network [@Ahuja1993] and apply push-relabel [@GoldbergTarjan1988],
cycle cancelling, network simplex, or Orlin's strongly polynomial algorithm
[@Orlin1993]. These are the most general formulations, which makes them the
right base for variable-ratio designs where a group absorbs several units. On a
plain one-to-one problem they are the slowest choice.

**Special-purpose routines** cover cases where the general machinery is wasted:
exhaustive enumeration below nine units, Hopcroft-Karp style bipartite matching
[@HopcroftKarp1973] when costs are binary or constant, Ramshaw-Tarjan
[@RamshawTarjan2012] for strongly rectangular problems where padding to a square
matrix would dominate the work, and bottleneck assignment, which minimises the
largest pair cost instead of the sum.

Two extensions return something other than a single optimal assignment. Murty's
ranking procedure [@Murty1968] with Lawler's partition scheme [@Lawler1972]
enumerates the k cheapest assignments, which lets a user see how much of the
total cost is forced and how much is arbitrary. Sinkhorn iteration [@Cuturi2013]
solves the entropy-regularised relaxation, returning a doubly stochastic
coupling rather than a permutation.

# A matching workflow

The package ships `hospital_staff`, a synthetic personnel dataset that supports
a treated/control workflow without external data. The `nurses_extended` table
holds 200 units and `controls_extended` holds 300, with age, experience,
hourly rate, department, certification level and a full-time indicator.

```{r workflow-data, echo=TRUE}
library(couplr)
data(hospital_staff)

treated <- transform(hospital_staff$nurses_extended,   id = nurse_id)
control <- transform(hospital_staff$controls_extended, id = nurse_id)
```

`match_couples()` is the front door. It takes the two data frames, the
covariates to match on, and returns the optimal one-to-one pairing under the
requested distance. Setting `auto_scale = TRUE` runs the preprocessing step:
covariates are screened for degenerate variance and excessive missingness, and a
scaling method is chosen so that variables measured in years and variables
measured in ordinal levels contribute comparably.

```{r workflow-match, echo=TRUE}
covars <- c("age", "experience_years", "certification_level")

m <- match_couples(
  left  = treated,
  right = control,
  vars  = covars,
  auto_scale = TRUE
)
m
```

The result is an S3 object of class `matching_result` with three components.
`m$pairs` is a tibble of matched pairs carrying the identifiers, the distance
for each pair and a signed per-variable difference column for every matching
variable, which makes it possible to see which covariate drives a poor pair
without recomputing anything. `m$unmatched` holds the identifiers left over on
each side, and `m$info` records the method used, the number of pairs and the
total distance.

Balance is the standard for judging whether a matched sample is usable.
`balance_diagnostics()` computes standardised mean differences, variance ratios
and Kolmogorov-Smirnov statistics on the matched sample, and `balance_table()`
formats them.

```{r workflow-balance, echo=TRUE}
bal <- balance_diagnostics(m, treated, control, vars = covars)
balance_table(bal)
```

Experience and certification fall below the conventional 0.1 threshold on the
absolute standardised difference [@Stuart2010], while age remains slightly
above it at 0.116; all three are below 0.15. `join_matched()` then returns
the analysis-ready frame, one row per pair, with every column of both input
tables carried through under `_left` and `_right` suffixes, so an outcome model
can be fitted directly on the matched data.

```{r workflow-join, echo=TRUE}
matched <- join_matched(m, treated, control)
dim(matched)
```

## From matched pairs to an estimate

The matched frame is the input to an outcome analysis rather than the end of
one. We take the hourly rate as the outcome. The paired difference within each
matched pair then estimates the association between full-time status and pay
among nurses comparable on age, experience and certification.

```{r workflow-effect, echo=TRUE}
d  <- matched$hourly_rate_left - matched$hourly_rate_right
tt <- t.test(d)
c(estimate = mean(d), lower = tt$conf.int[1], upper = tt$conf.int[2])
```

Full-time nurses earn `r sprintf("%.2f", mean(matched$hourly_rate_left - matched$hourly_rate_right))`
more per hour than their matched part-time counterparts, with a 95% interval
that excludes zero. This is a paired difference on a matched sample and rests
on assumptions no matching method can verify: conditional ignorability,
positivity, and the stable unit treatment value assumption (SUTVA), under which
a unit's outcome depends on its own treatment alone. The matched output can
also be passed to an outcome model for regression adjustment; valid estimation
and uncertainty quantification still depend on the estimator used [@Stuart2010].

What matching can offer against the first of those assumptions is a statement
about how strong an unmeasured confounder would have to be to overturn the
result. `sensitivity_analysis()` implements Rosenbaum bounds for the Wilcoxon
signed-rank statistic on one-to-one matched pairs, reporting upper and lower
p-value bounds across a grid of the odds-ratio parameter $\Gamma$
[@Rosenbaum2002].

```{r workflow-sensitivity, echo=TRUE}
sens <- sensitivity_analysis(m, treated, control, outcome_var = "hourly_rate")
sens$results[sens$results$gamma %in% c(1, 1.5, 1.75, 2), ]
```

At $\Gamma = 1$, meaning no hidden bias, the association is significant. It
survives an unmeasured confounder that raises one pair member's odds of being
full-time by half, $\Gamma = 1.5$, and stops being significant at
$\Gamma =$ `r sens$critical_gamma`. A reader can judge from that number whether
a confounder of that strength is plausible in this setting; the analysis
quantifies exposure to hidden bias rather than removing it.

# Constraints, designs and interoperability

## Reshaping the cost matrix

Three mechanisms restrict which pairs are admissible, and all three work by
writing non-finite entries into the cost matrix before it reaches the solver.
A global `max_distance` forbids any pair beyond a distance ceiling. Per-variable
`calipers` forbid pairs whose raw difference on a named variable exceeds a
threshold, independently of the distance metric in use. Explicit forbidden pairs
are passed as non-finite entries by the user. Every solver recognises a
non-finite entry as a missing edge, so no returned pair ever violates a
constraint.

Tight constraints usually make \@ref(eq:lap) infeasible, and the interesting
question is what happens then. couplr first drops the rows and columns whose
every edge is forbidden, which are reported unmatched. The remaining submatrix
can still fail Hall's condition, with several rows competing for the same few
admissible columns, and no assignment reaching every row. Rather than return
whichever pairs a heuristic happens to find, couplr replaces the forbidden
entries with a finite sentinel and re-solves with the same optimal solver. The
sentinel is set above $(k+1)$ times the spread of the admissible costs, where
$k$ is the largest possible number of pairs, so one sentinel edge outweighs any
saving available on the real edges. Minimising cost on the padded problem
therefore minimises the number of sentinel edges first and the real cost
second, and dropping the sentinel pairs leaves exactly the lexicographic
optimum: the largest admissible matching, cheapest among those of its size.

```{r constraints, echo=TRUE}
m_cal <- match_couples(
  treated, control, vars = covars,
  auto_scale   = TRUE,
  calipers     = list(age = 3, experience_years = 2),
  max_distance = 1.5
)
c(pairs = nrow(m_cal$pairs), unmatched_left = length(m_cal$unmatched$left))
```

The caliper costs pairs: `r nrow(m_cal$pairs)` of the 200 treated units keep a
partner. This is the intended behaviour of a caliper and the reason the
unmatched identifiers are returned rather than dropped, since the units that
fail to match define the population the estimate actually applies to. The
constraints here are severe, leaving 2,173 admissible pairs out of 60,000, so
the padded solve is what produces this result; a greedy pass over the same
admissible edges matches 180 units rather than
`r nrow(m_cal$pairs)`.

Blocking is the other structural constraint. `block_id` restricts matching to
run within levels of a factor, and `matchmaker()` constructs blocks when no
grouping variable exists, either from an existing variable or by k-means or
hierarchical clustering on chosen block variables. The solver is called once per
block, and blocked designs report per-block balance so that imbalance inside one
stratum is not averaged away across the others.

## Matching designs

Beyond one-to-one matching, `match_couples()` takes `ratio` for k-to-one designs
and `replace` for matching with replacement. Five further designs have their own
entry points, each returning an object in the same family:

- `ps_match()` estimates a propensity score [@RosenbaumRubin1983] and matches on
  it, with the caliper expressed in standard deviations of the linear predictor.
- `full_match()` assigns every unit to a matched group of variable size, so no
  unit is discarded. The optimal path solves a minimum-cost flow problem using
  Dijkstra's algorithm with Johnson potentials [@Johnson1977]; a greedy two-pass
  heuristic is available for large problems.
- `cem_match()` implements coarsened exact matching [@IacusKingPorro2012],
  binning covariates and matching exactly on the binned strata.
- `subclass_match()` divides the sample into propensity-score subclasses and
  returns subclass weights for estimating the average treatment effect on the
  treated (ATT) or the average treatment effect (ATE).
- `cardinality_match()` targets the cardinality-matching objective
  [@Zubizarreta2014], the largest matched sample meeting prespecified balance
  constraints. The implementation is an iterative pruning heuristic rather than
  the mixed-integer program: it starts from a full match and repeatedly removes
  the pairs contributing most to the worst-balanced variable until every
  standardised difference falls under the threshold. It therefore approximates
  the cardinality-matching solution and does not certify maximality.

## Interoperability

couplr fits into a workflow built around the established packages.
`as_matchit()` converts a couplr result into a `matchit` object, so downstream
code written against \pkg{MatchIt} keeps working, and `match_data()` returns
data in the layout that \pkg{MatchIt} users expect. `bal.tab()` methods are
registered for the couplr result classes, so \CRANpkg{cobalt} produces its usual
balance output on a couplr match. Weighted matched data flows into
\CRANpkg{marginaleffects} [@marginaleffects] for estimation.

# The solver layer

The solver layer is reachable on its own, for a user who arrives with a cost
matrix in hand. `lap_solve()` and `assignment()` take a matrix and return the
optimal assignment; `lap_solve()` also accepts a long data frame of source,
target and cost columns, which is the natural shape when costs arrive from a
database.

```{r solver-basic, echo=TRUE}
set.seed(1)
cost <- matrix(sample.int(100, 25, replace = TRUE), nrow = 5)
res  <- lap_solve(cost)
res
```

The `method` argument names any of the nineteen solvers; `"auto"` is the
default. Dual potentials come from `assignment_duals()`, which returns the
vectors $u$ and $v$ of \@ref(eq:duals) alongside the assignment, so all four
parts of the certificate can be checked independently of the solver:

```{r solver-duals, echo=TRUE}
d   <- assignment_duals(cost)
i   <- which(d$match > 0)
j   <- d$match[i]
sel <- cbind(i, j)

c(primal   = length(i) == min(dim(cost)) && !anyDuplicated(j),
  dual     = all(outer(d$u, d$v, "+") <= cost + 1e-9),
  tight    = all(abs(d$u[i] + d$v[j] - cost[sel]) < 1e-9),
  no_gap   = abs(sum(cost[sel]) - sum(d$u) - sum(d$v)) < 1e-9)
```

Four extensions operate on the same matrix. `lap_solve_kbest()` returns the k
cheapest assignments in increasing order of cost, which shows how much of a
solution is forced by the data. `bottleneck_assignment()` minimises the largest
individual pair cost, the right objective when the worst pair rather than the
average pair is what matters. `lap_solve_batch()` solves a stack of independent
problems, optionally across threads. `sinkhorn()` returns the entropy-regularised
coupling, and `sinkhorn_to_assignment()` rounds it to a permutation.

```{r solver-kbest, echo=TRUE}
kb <- lap_solve_kbest(cost, k = 3)
tapply(kb$total_cost, kb$rank, unique)
```

The two cheapest assignments both cost 120 and the third costs 150, so the
optimum here is attained by two distinct pairings. A matching study reading the
same output would learn that its matched sample is not unique, which stays
invisible when a solver returns one assignment.

# Software design

## Layers and object model

The package is organised in three layers. The first converts user input into a
cost matrix: it validates the data frames, screens covariates, applies scaling
and weights, computes distances, and writes constraints into the matrix. The
second solves \@ref(eq:lap). The third builds result objects and the methods
that consume them.

The object model is S3 throughout. Matching results carry the class vector
`c("matching_result", "couplr_result")`, with the design-specific classes
(`full_matching_result`, `cem_result`, `subclass_result`) sharing the
`couplr_result` parent. The shared parent is what makes the diagnostic and
conversion layer generic: `balance_diagnostics()`, `join_matched()`,
`match_data()` and the `bal.tab()` methods dispatch on it once rather than being
rewritten per design. We chose S3 deliberately. S4 or R6 would have bought
encapsulation the package does not need, at the cost of making result objects
harder to inspect interactively and harder to serialise. Results are plain
lists of tibbles, so a user who wants something the package does not provide
can reach in and take it.

Solver results are a separate family. `lap_solve()` returns a tibble subclassed
as `lap_solve_result` with `source`, `target` and `cost` columns, which means the
optimum is immediately usable in \CRANpkg{ggplot2} [@ggplot2] or a join without
an extraction step, while `get_total_cost()` and `get_method_used()` recover the
metadata carried in attributes.

## C++ organisation

All nineteen solvers are written from scratch in C++ and exposed through
\CRANpkg{Rcpp} [@Rcpp]. The `src` tree separates concerns: `src/core` holds the
cost-source types and shared utilities with no Rcpp dependency, `src/solvers`
holds one algorithm per file with a thin `_rcpp.cpp` wrapper each, and
`src/interface` holds the conversion between R objects and the internal types.
The separation means the solver bodies are ordinary C++ that can be compiled and
tested outside R.

We template solvers on their cost source rather than writing them against a
concrete matrix. A cost source has to provide `at(i, j)`, `allowed(i, j)` and
`empty()`. Two types
satisfy this interface: `CostMatrix`, which owns a dense flat array plus a mask,
and `LazyCostMatrix`, which owns the two feature matrices and computes
$C_{ij}$ on demand from the requested metric, applying calipers and the distance
ceiling in `allowed()`. A handful of hot loops index the dense mask array
directly for speed, which the lazy type cannot support; these are guarded by a
`supports_raw_mask` trait and an `if constexpr` branch rather than by duplicating
the solver. One solver body therefore serves both cost representations, and
there is no second implementation to keep in step.

`LazyCostMatrix` also bakes in the transformations that the dense path applies
in a separate preparation pass, namely mapping forbidden entries to a large
sentinel and negating costs when maximising. Doing this at construction is what
allows the untouched solver body to work on either type.

## Automatic solver selection

`method = "auto"` inspects the problem and picks a solver. The rules are few,
and the benchmark in the next section supports rule 5 directly. Rules 1 to 4
are initial heuristics: they follow from the structure each special-purpose
routine exploits, and the thresholds in them are not calibrated against a
benchmark that varies sparsity or rectangularity.

1. Problems with at most eight rows and eight columns go to exhaustive
   enumeration, which is exact and faster than setting up a general solver.
2. Cost matrices whose finite entries are constant or lie in $\{0, 1\}$ go to the
   Hopcroft-Karp style routine, which exploits the absence of a real cost scale.
3. Matrices with more than half of their entries non-finite go to `lapmod`, the
   sparse Jonker-Volgenant variant, which carries forbidden edges in its
   adjacency structure at every size.
4. Strongly rectangular problems, with at least three times as many columns as
   rows, go to shortest augmenting paths, avoiding the padding a square-oriented
   solver would need.
5. Everything else goes to Jonker-Volgenant.

Rectangular problems are transposed internally so that the solver always sees at
least as many columns as rows, and the assignment is mapped back afterwards; the
user never sees the transpose.

Rules 2 and 3 read the entries, and so does the rejection of `NaN` inputs that
runs for every method, so choosing a solver means a pass over the cost matrix.
Expressing that pass in R cost an allocation the size of the matrix for each
test, through `any(is.nan())`, `range(finite = TRUE)` and
`mean(is.na() | is.infinite())`, and the overhead made `"auto"` up to twice as
slow as naming the solver it would pick. A single C++ pass now returns every
quantity the rules need, with no allocation and no second scan: the binary-cost
test is a branch inside the same loop rather than a `unique()` over the finite
entries, and an integer cost matrix is read as integers rather than coerced.
That leaves the inspection a linear read in front of a superlinear solve, so its
share of the runtime falls as problems grow, and the dashed line in
Figure \@ref(fig:solver-bench) sits on the solver the dispatcher selects. The
pass is skippable for callers who already know the shape of their problem, by
naming the method.

## Memory modes

The dense representation of the cost matrix is the binding constraint on problem
size long before runtime is. A 100,000 by 100,000 problem needs 80 GB as doubles;
the same units described by 100 covariates need 80 MB. couplr exposes this as
`memory_mode`, with values `"dense"`, `"lazy"` and `"auto"`.

Under `"lazy"`, no cost matrix is built. The solver holds the feature matrices
and evaluates distances as it requests them, which trades arithmetic for
memory and returns the same assignment the dense path returns. The lazy path
covers Jonker-Volgenant and auction with the built-in metrics, together with
calipers and the distance ceiling.

Under `"auto"`, the package decides. Below ten million cells, about 80 MB dense,
it skips the check entirely, since probing free memory on every ordinary problem
would be pure overhead. Above that, we estimate the dense footprint with a
factor of four over the raw eight bytes per cell, which covers the copies made
during conversion and preparation that can coexist under garbage-collection
lag, and compare that against available system memory. Available memory is read
per platform, from `MemAvailable` on Linux, from `vm_stat` page counts on macOS,
including inactive and speculative pages because the kernel keeps almost nothing
on the free list, and from `Win32_OperatingSystem` on Windows. If the estimate
exceeds half of what is available, `"auto"` switches to the lazy path where one
exists and warns explicitly where one does not, naming blocking, greedy matching
or a smaller problem as the alternatives. Detection failure is treated as
unknown rather than as success: a fixed 4 GB threshold takes over, so the guard
never silently disappears on an unrecognised platform.

## Verification

Statistical smoke tests establish that a function returns an object of the right
shape. For a solver, the corresponding standard is whether the returned
assignment is the optimal one. We check every optimal solver against
exhaustive enumeration on randomised instances covering integer and fractional
costs, square and rectangular shapes, minimisation and maximisation, and
matrices containing forbidden edges. The brute-force reference is the same
`bruteforce` solver the dispatcher uses below nine units, so the verified path
and the shipped path are one implementation.

Constrained matching is checked against the same standard, and needs its own
reference because the objective is lexicographic. An enumeration over all
partial matchings of a small instance gives the largest admissible matching and
the cheapest total among matchings of that size, and the constrained solve has
to reproduce both. That test is what showed us the earlier behaviour was wrong:
until version 1.5.6, a constrained problem whose admissible graph had no
complete matching returned a greedy partial matching under the name of the
optimal solver that had been requested. The package carries 126 test files in
total, covering the R layer, the C++ wrappers, dispatch rules and the matching
designs.

Dual feasibility gives a second, independent check that a user can run on their
own problem, since \@ref(eq:duals) can be verified in $O(nm)$ without solving
anything, as shown in the solver section above.

## Dependency footprint

couplr declares \CRANpkg{Rcpp}, \CRANpkg{tibble} [@tibble], \CRANpkg{dplyr},
\CRANpkg{rlang}, \CRANpkg{purrr} and `methods` in `Imports`, and `Rcpp` alone in
`LinkingTo`. The recursive closure of hard dependencies is 17 non-base packages,
so a default installation brings 18 packages including couplr itself. The
comparable figures on the same day were 18 for \pkg{optmatch}, 8 for
\pkg{MatchIt} and 7 for \pkg{designmatch}.

Two choices sit behind that number. First, we use no external solver: all
nineteen algorithms are package source, so there is no dependency on an LP or
MIP library and no system solver to install. Second, everything that serves a
single feature lives in `Suggests` and is loaded at call time. The animation and image-morphing functions, the plotting
helpers, the interoperability layer for \pkg{MatchIt}, \pkg{optmatch} and
\pkg{cobalt}, and the parallel back end are all optional in this sense, so a user
who only wants to match pays for none of them. `LinkingTo` deserves particular
care, since every entry there forces a compile of that package before couplr can
be built from source; couplr declares only what its headers actually include.

# Performance

## Solver benchmark

Figure \@ref(fig:solver-bench) reports median wall-clock solve time against
problem size for all nineteen solvers, grouped by family. We draw costs as
integers uniformly from $[1, 10{,}000]$ on square matrices and report medians of
five replicates on a single core of an Apple M4 Pro. The slower families are
capped at smaller sizes so that the scalable solvers can be measured to
$n = 5000$ within a reasonable budget. All timings in this section were measured
on 2026-08-09 under R 4.6.0 with couplr `r couplr_ver`, \pkg{optmatch} 0.10.8
and \pkg{MatchIt} 4.7.2.

```{r solver-bench, fig.height=6.2, fig.cap="Median wall-clock solve time against problem size for the nineteen assignment solvers in couplr, grouped by algorithm family on shared log-log axes. Seventeen solvers are timed on square integer cost matrices with entries drawn uniformly from 1 to 10,000. The two special-purpose solvers in the Other panel run on their own inputs and are not comparable to the rest: HK-01 is timed on binary cost matrices, and Brute-F only up to n = 8. The dashed grey line repeated in every panel is the automatic dispatcher on the uniform integer costs. Medians of five replicates on a single core of an Apple M4 Pro. Cubic-time and general flow solvers are capped at smaller sizes, which is why their lines end early.", fig.alt="Five panels of log-log line plots showing solve time in milliseconds against problem size n from 4 to 5000. Solve time rises with problem size in every panel. The Jonker-Volgenant panel reaches n equals 5000 in roughly one second, the fastest of the families. Auction, cost-scaling and flow-based solvers are one to three orders of magnitude slower at matched sizes and stop at n equals 1000 or 2000. The dashed dispatcher line lies on top of the fastest solver in each panel."}
fam_map <- c(
  "JV / Augmenting path"  = "JV / augmenting path",
  "Auction"               = "Auction",
  "Cost scaling"          = "Cost scaling",
  "Flow-based"            = "Flow-based",
  "Classical"             = "Other",
  "Rectangular / general" = "Other",
  "Special-purpose"       = "Other"
)

panel_levels <- c("JV / augmenting path", "Auction", "Cost scaling",
                  "Flow-based", "Other")

solvers <- bench[bench$family != "Auto", ]
solvers$panel <- factor(fam_map[solvers$family], levels = panel_levels)

auto <- bench[bench$family == "Auto", ]
auto_all <- do.call(rbind, lapply(panel_levels, function(p) {
  transform(auto, panel = factor(p, levels = panel_levels))
}))

pal <- c(
  "JV" = "#0072B2", "SAP" = "#56B4E9", "LAPMOD" = "#004E73",
  "Auction" = "#E69F00", "Auction-GS" = "#8C5200", "Auction-S" = "#F0C066",
  "CSA" = "#009E73", "Gabow-T" = "#004534", "SSAP-B" = "#66C2A5",
  "CS-flow" = "#CC79A7", "Cycle-C" = "#7B2D5E", "Push-R" = "#E5B8D4",
  "Net-S" = "#6A3D9A", "Orlin" = "#B07AA1",
  "Hungarian" = "#D55E00", "Munkres" = "#8C3B00", "Ramshaw" = "#5A4634",
  "HK-01" = "#117733", "Brute-F" = "#999933"
)

# Curves run from the lower left to the upper right on log-log axes, so the
# upper left of every panel is free for an in-panel key.
key_order <- list(
  "JV / augmenting path" = c("JV", "SAP", "LAPMOD"),
  "Auction"              = c("Auction", "Auction-GS", "Auction-S"),
  "Cost scaling"         = c("CSA", "Gabow-T", "SSAP-B"),
  "Flow-based"           = c("CS-flow", "Cycle-C", "Push-R", "Net-S", "Orlin"),
  "Other"                = c("Hungarian", "Munkres", "Ramshaw", "HK-01",
                             "Brute-F")
)
lt_pool <- c("solid", "22", "12", "4212", "62")
lt <- unlist(lapply(key_order, function(v) setNames(lt_pool[seq_along(v)], v)))
names(lt) <- unlist(key_order)

key <- do.call(rbind, lapply(names(key_order), function(p) {
  data.frame(panel = factor(p, levels = panel_levels),
             label = key_order[[p]],
             k = seq_along(key_order[[p]]) - 1,
             stringsAsFactors = FALSE)
}))
x_lim <- c(3.5, 6000)
y_lim <- c(0.008, 2e5)
lx <- log10(x_lim); ly <- log10(y_lim)
key$x    <- 10^(lx[1] + 0.05 * diff(lx))
key$xend <- 10^(lx[1] + 0.17 * diff(lx))
key$y    <- 10^(ly[2] - (0.05 + 0.085 * key$k) * diff(ly))

ggplot(solvers, aes(n, median_ms, colour = label, linetype = label)) +
  geom_line(data = auto_all, aes(n, median_ms), inherit.aes = FALSE,
            colour = "grey45", linetype = "dashed", linewidth = 0.5) +
  geom_line(linewidth = 0.6) +
  geom_segment(data = key, aes(x = x, xend = xend, y = y, yend = y,
                               colour = label, linetype = label),
               linewidth = 0.6, inherit.aes = FALSE) +
  geom_text(data = key, aes(x = xend, y = y, label = label),
            hjust = -0.15, vjust = 0.45, size = 3.1, colour = "grey15",
            inherit.aes = FALSE) +
  facet_wrap(~ panel, ncol = 2) +
  scale_x_log10(breaks = c(10, 100, 1000), limits = x_lim,
                expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_log10(breaks = c(0.01, 1, 100, 10000),
                labels = c("0.01 ms", "1 ms", "100 ms", "10 s"),
                limits = y_lim, expand = expansion(mult = c(0.02, 0.02))) +
  scale_colour_manual(values = pal, guide = "none") +
  scale_linetype_manual(values = lt, guide = "none") +
  labs(x = "problem size, n", y = "median solve time") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
    strip.text = element_text(face = "bold", hjust = 0, size = 11),
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 11),
    axis.line = element_line(colour = "grey20", linewidth = 0.4)
  )
```

Jonker-Volgenant with its warm start is the fastest general-purpose solver at
every measured size, which is why it is the dispatcher's default. The auction
and cost-scaling families are competitive at small sizes and fall behind on
dense uniform costs, where their advantage, tolerance of degenerate structure
and a better asymptotic bound in the Gabow-Tarjan case, does not pay. General
flow formulations are the slowest, as expected for machinery that solves a
larger problem than the one posed. The two special-purpose routines run on the
inputs they exist for, exhaustive enumeration below nine units and the
Hopcroft-Karp style solver on binary costs, so their timings are not comparable
with the rest of the figure and are shown only to establish that each is
practical on its own input.

These orderings hold for dense uniform integer costs on square matrices, and
that is the whole of what the figure establishes. They fix the dispatcher's
default. The remaining rules rest on the structure the special-purpose routines
exploit rather than on measurement, and the regimes that would test them,
sparse, strongly rectangular and degenerate cost scales, are left to the
benchmark extension described in the summary.

## Balance against established alternatives

Speed only matters once the packages agree on the answer. We fix the task in
Figure \@ref(fig:love), one-to-one optimal
matching on a Mahalanobis distance with pooled within-group covariance, and run
it on the LaLonde NSW data [@LaLonde1986; @DehejiaWahba1999], 185 treated units,
429 controls and eight covariates. The pooled within-group convention is the one
`optmatch::match_on()` uses by default and the one couplr adopts.

```{r love, fig.height=3.4, fig.cap="Absolute standardised mean differences on the eight LaLonde NSW covariates, before and after one-to-one optimal Mahalanobis matching with pooled within-group covariance. couplr, MatchIt and optmatch agree to three decimal places after matching, so a single matched point is shown per covariate. The dashed line marks the conventional 0.1 threshold. The indicator for Black respondents starts far outside the plotted range at 1.757 and remains at 1.053 after matching, a residual imbalance that no one-to-one matcher can resolve on this data.", fig.alt="A dot plot with eight covariates on the vertical axis and absolute standardised mean difference on the horizontal axis. For each covariate an open circle marks the value before matching and a filled square the value after matching, joined by a grey line. Every covariate moves left toward zero. Five of the eight land below the dashed 0.1 threshold, married and 1974 earnings sit just above it near 0.12 and 0.13, and the indicator for Black respondents remains far to the right at 1.05."}
pretty <- c(age = "age", educ = "education (years)",
            race_Black = "race: Black", race_Hispanic = "race: Hispanic",
            married = "married", nodegree = "no high-school degree",
            re74 = "earnings, 1974", re75 = "earnings, 1975")
lp <- lalonde
lp$lab <- pretty[lp$covar]
lp <- lp[order(lp$unmatched), ]
lp$lab <- factor(lp$lab, levels = lp$lab)
lp$matched <- lp$couplr

long <- rbind(
  data.frame(lab = lp$lab, state = "before matching", smd = lp$unmatched),
  data.frame(lab = lp$lab, state = "after matching",  smd = lp$matched)
)
long$state <- factor(long$state, levels = c("before matching", "after matching"))

ggplot(long, aes(smd, lab, colour = state, shape = state)) +
  geom_segment(data = lp, aes(x = unmatched, xend = matched, y = lab, yend = lab),
               inherit.aes = FALSE, colour = "grey70", linewidth = 0.4) +
  geom_vline(xintercept = 0.1, linetype = "22", colour = "grey30",
             linewidth = 0.4) +
  geom_point(size = 2.4, stroke = 0.9) +
  scale_colour_manual(values = c("before matching" = "#D55E00",
                                 "after matching"  = "#0072B2"), name = NULL) +
  scale_shape_manual(values = c("before matching" = 1,
                                "after matching"  = 15), name = NULL) +
  scale_x_continuous(limits = c(-0.02, 1.85),
                     breaks = c(0, 0.1, 0.5, 1.0, 1.5)) +
  labs(x = "|standardised mean difference|", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3),
    legend.position = "bottom",
    axis.line.x = element_line(colour = "grey20", linewidth = 0.4)
  )
```

The three packages produce identical absolute standardised mean differences to
three decimal places, so the matched points coincide and one is plotted.
Matching marginals could still hide different assignments, so we score all
three on one objective: the same Mahalanobis distance matrix, summed over each
package's chosen pairs. Each matches all
`r lal_obj$n_pairs[lal_obj$package == "couplr"]` treated units, at a total
distance of `r sprintf("%.3f", lal_obj$total_cost[lal_obj$package == "couplr"])`
for couplr against
`r sprintf("%.3f", lal_obj$total_cost[lal_obj$package == "optmatch"])` for both
alternatives. They agree on the optimum, not only on its marginal summaries.

Five of the eight covariates fall below 0.1 after matching; `married` and 1974
earnings sit just above at 0.124 and 0.128, and the indicator for Black
respondents drops from 1.757 to 1.053 without approaching the threshold, which
is a property of this dataset rather than of any package: the control pool does
not contain enough Black respondents for a one-to-one match to balance that
indicator.

## Scaling

Table \@ref(tab:scaling) reports median wall-clock time on synthetic problems
with the same eight-covariate structure, at six sizes with a one-to-two ratio of
treated to control units. Every package materialises the distance matrix here,
so we time couplr with `memory_mode = "dense"`.

```{r scaling, echo=FALSE}
lab <- c("167 + 333", "667 + 1,333", "1,667 + 3,333",
         "3,333 + 6,667", "6,667 + 13,333", "16,667 + 33,333")
sizes <- c(500, 2000, 5000, 10000, 20000, 50000)

fmt <- function(pkg) {
  vapply(sizes, function(n) {
    r <- scaling[scaling$n_total == n & scaling$package == pkg, ]
    if (nrow(r) == 0) return("not run")
    if (is.na(r$median_s[1])) {
      return(if (grepl("timeout", r$status[1])) "timeout" else "int overflow")
    }
    s <- r$median_s[1]
    if (s < 1) sprintf("%.0f ms", s * 1000) else sprintf("%.3g s", s)
  }, character(1))
}

tab <- data.frame(
  `Problem size` = lab,
  couplr   = fmt("couplr"),
  optmatch = fmt("optmatch"),
  MatchIt  = fmt("MatchIt"),
  check.names = FALSE
)

knitr::kable(
  tab, align = "lrrr", booktabs = TRUE,
  caption = "One-to-one optimal Mahalanobis matching, median wall-clock time by problem size. Treated to control ratio 1:2, eight covariates, pooled within-group covariance, single core with single-threaded BLAS on an Apple M4 Pro. Medians of 5, 5, 3, 3, 1 and 1 replicates for the six rows. The optmatch option max.problem.size was set to Inf for the two largest sizes. int overflow marks an integer-size overflow inside the optmatch back end reached through MatchIt; timeout marks a replicate exceeding the 300-second cap."
)
```

The margin widens with problem size. couplr is roughly eight times faster than
\pkg{optmatch} at $n = 500$ and twenty-seven times faster at $n = 20{,}000$,
where it finishes in
`r sprintf("%.1f", scaling$median_s[scaling$n_total == 20000 & scaling$package == "couplr"])`
seconds against 4.7 minutes. \pkg{MatchIt} runs
between ten and twenty-five times slower up to $n = 10{,}000$, which is the
largest size it completes before `matchit(method = "optimal")` aborts inside the
\pkg{optmatch} back end with an integer-size overflow. At $n = 50{,}000$ couplr
finishes in
`r sprintf("%.0f", scaling$median_s[scaling$n_total == 50000 & scaling$package == "couplr"])`
seconds and both alternatives exceed the 300-second per-replicate cap.

A likely contributor is the difference in formulation rather than a faster inner
loop in any single algorithm. \pkg{optmatch} solves a minimum-cost flow problem,
which is the right formulation for the full and variable-ratio designs it is
built around and more machinery than a plain one-to-one problem needs; couplr's
dispatcher sends this dense square problem to Jonker-Volgenant. Separating that
contribution from implementation differences would need a benchmark that holds
the formulation fixed across packages, which these timings do not do.

The lazy path removes the memory ceiling as well. The same $n = 50{,}000$ match
under `memory_mode = "lazy"` completes in
`r sprintf("%.0f", lazy$elapsed_s[lazy$n_total == 50000])` seconds without ever
building the distance matrix, returning the assignment the dense path returns.
At $n = 20{,}000$ it takes
`r sprintf("%.1f", lazy$elapsed_s[lazy$n_total == 20000])` seconds against
`r sprintf("%.1f", scaling$median_s[scaling$n_total == 20000 & scaling$package == "couplr"])`
for the dense path, so on these problems recomputing distances is cheaper than
materialising and traversing the matrix.

```{=latex}
\FloatBarrier
```

## Feature coverage

Table \@ref(tab:capability) lists the features that separate the three packages.
Rows where all three offer the feature, covariate distance, propensity-score
matching, exact blocking and a cost-matrix entry point, are omitted.

```{r capability, echo=FALSE}
cap <- data.frame(
  Feature = c("Per-variable calipers from a named vector",
              "Rectangular problems without padding",
              "Sparse cost support",
              "k-best assignments",
              "Bottleneck (minimax) assignment",
              "Rosenbaum sensitivity bounds",
              "User-selectable assignment algorithm"),
  couplr = c("yes", "yes", "yes", "yes", "yes", "yes", "19 solvers"),
  MatchIt = c("yes", "via optmatch", "no", "no", "no", "no", "via optmatch"),
  optmatch = c("partial", "via padding", "yes", "no", "no", "no", "2 back ends"),
  check.names = FALSE
)
knitr::kable(
  cap, align = "lccc", booktabs = TRUE,
  caption = "Selected assignment-layer features. The table covers the layer this article is about and deliberately omits areas where the alternatives lead, among them the breadth of matching designs reachable through a single MatchIt call and the maturity of optmatch's full-matching implementation. Row meanings: per-variable calipers are marked partial for optmatch because its calipers threshold one distance object at a time, so a per-variable set requires combining objects; rectangular problems are accepted by both alternatives as unequal group sizes, but reach the solver as a padded or variable-ratio formulation rather than as a rectangular cost matrix; the assignment algorithm is user-selectable in optmatch through solver = (RELAX-IV or LEMON, with a choice of LEMON algorithm) and in MatchIt by forwarding solver to optmatch::fullmatch(). Assessed 2026-08-09 against MatchIt 4.7.2 and optmatch 0.10.8."
)
```

# Summary and discussion

We built couplr on the observation that one optimisation problem sits underneath
a set of tasks that R currently treats as unrelated. Making the assignment
problem the organising object, rather than a hidden step inside a matching
function, is what lets a single package serve a causal-inference user starting
from covariates and an operations-research user starting from a cost matrix.
Three design consequences follow from that decision. The object model gives the
result classes a shared parent, so diagnostics are written once. Solvers are
templated on their cost source, so the dense and lazy paths cannot drift apart.
A dispatcher turns the choice among nineteen solvers into an implementation
detail for users who would rather not make it.

couplr has four limitations. The lazy cost path covers Jonker-Volgenant and
auction with the built-in metrics; other solvers and
user-supplied distance functions still require a dense matrix, and `full_match()`
uses a flow back end that has no lazy path yet. `cardinality_match()` is a
pruning heuristic and does not certify that the matched sample it returns is the
largest one satisfying the balance constraints, which the mixed-integer
formulation would. The dispatch rules are calibrated on dense uniform integer
costs and on the sparsity and rectangularity boundaries that follow from them;
cost structures far from that regime may be better served by naming a solver
explicitly, which is why every solver stays reachable by name. Matching on
observed covariates cannot address unmeasured confounding, and the sensitivity
analysis quantifies exposure to it rather than removing it.

We are extending the lazy representation to more solvers and adding a
minimum-cost flow path for variable-ratio designs under the same templated cost
interface. We are also widening the benchmark beyond uniform integer costs, so
that the dispatch rules can be recalibrated on sparse, rectangular and
clustered cost matrices rather than resting on the structure each routine
exploits.

couplr is available on CRAN and developed at
`https://github.com/gcol33/couplr`, with documentation at
`https://gillescolling.com/couplr/`. Bug reports and feature requests go through
the GitHub issue tracker.

# Acknowledgements

No external funding supported this work.
